### 2024-03-26 paper/paper.qmd
updated README and paper

---
title: "An Investigation Into The Impact of Transit Improvements on Toronto Traffic Volumes 2010 - 2019"
subtitle: "My subtitle if needed"
author: 
  - Maria Mangru
thanks: "Code and data are available at: https://github.com/MariaMangru/An-Investigation-Into-The-Impact-of-Transit-Improvements-on-Toronto-Traffic-Volumes-2010-2019."
date: today
date-format: long
abstract: "In this study, I have embarked on an analysis of a decade's worth of traffic volume data from critical intersections in Toronto, scrutinizing the effects of public transit developments on urban traffic congestion from 2010 to 2019. [INSERT FINDINGS HERE] The significance of this research lies in its potential to inform and shape urban infrastructure strategies, offering evidence on how transit policies can alleviate traffic pressures in metropolitian areas."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(here)
```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/analysis_data/cleaned-data-2010-2019.csv"))

```

Talk way more about it.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

\newpage

# References

### 2024-03-26 scripts/05-model.R

@@ -13,7 +13,7 @@ library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+analysis_data <- read_csv("data/analysis_data/cleaned-data-2010-2019.csv")
 
 ### Model data ####
 first_model <-

### 2024-03-26 scripts/05-model.R
update project name

@@ -1,16 +1,14 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Creates negative binomial model 
+# Author: Maria Mangru
+# Date: March 2024
+# Contact:maria.mangru@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
 
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+
 
 #### Read data ####
 analysis_data <- read_csv("data/analysis_data/cleaned-data-2010-2019.csv")

### 2024-03-30 scripts/02-data_cleaning.R
filtered dataset to focus on Spadina Expansion
Code still needs to be cleaned to avoid repeated steps.

@@ -8,33 +8,38 @@
 
 #### Workspace setup ####
 library(dplyr)
+library(stringr)
 
 #### Clean data ####
+colnames(data)
 
 # access raw data 
 data <- read.csv("./data/raw_data/raw-data-2010-2019.csv")
 
 # Removing less relevant columns to focus on total traffic volumes, buses, pedestrians, and cyclists
 data_simplified <- data[, c("count_id", "count_date", "location_id", "location", 
-                            "sb_cars_t", "nb_cars_t", "wb_cars_t", "eb_cars_t",
-                            "sb_bus_r", "nb_bus_r", "wb_bus_r", "eb_bus_r", 
-                            "nx_peds", "sx_peds", "ex_peds", "wx_peds", 
-                            "nx_bike", "sx_bike", "ex_bike", "wx_bike")]
+                            "sb_cars_r","sb_cars_t","sb_cars_l","nb_cars_r", "nb_cars_t", "nb_cars_l", "wb_cars_r",
+                            "wb_cars_t","wb_cars_l", "eb_cars_r", "eb_cars_t","eb_cars_l", "sb_bus_r", "sb_bus_t", 
+                            "sb_bus_l", "nb_bus_r", "nb_bus_t", "nb_bus_l", "wb_bus_r", "wb_bus_t", "wb_bus_l", "eb_bus_r", "eb_bus_t", 
+                            "eb_bus_l", "nx_peds", "sx_peds", "ex_peds", "wx_peds", "nx_bike", "sx_bike", "ex_bike", "wx_bike", "centreline_type")]
 
-# Correct data format
-# data_simplified$count_date <- as.Date(data_simplified$count_date, format = "%m/%d/%Y")
+# Remove centreline_type = 1
+#data_simplified <- data_simplified %>%
+  #filter(centreline_type != 1)
+
+
 
 # Aggregating data to daily traffic volumes 
 daily_traffic_modes <- data_simplified %>%
   group_by(count_date, location_id, location) %>%
-  summarise(daily_sb_cars_t = sum(sb_cars_t, na.rm = TRUE),
-            daily_nb_cars_t = sum(nb_cars_t, na.rm = TRUE),
-            daily_wb_cars_t = sum(wb_cars_t, na.rm = TRUE),
[…]

### 2024-03-30 scripts/05-model.R

@@ -6,30 +6,27 @@
 # License: MIT
 
 #### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
+library(MASS) 
+library(dplyr)
+library(readr)
 
+
+## NEEDS A LOT OF WORK
 
 #### Read data ####
 analysis_data <- read_csv("data/analysis_data/cleaned-data-2010-2019.csv")
 
 ### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
+traffic_model_nb <- glm.nb(
+  formula = daily_sb_cars_t ~ daily_sb_bus + daily_wb_bus,
+  data = analysis_data,
+  link = log
+)
 
+summary(traffic_model_nb)
 
 #### Save model ####
 saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+  traffic_model_nb,
+  file = "model/traffic_model_nb.rds"
 )
-
-

### 2024-03-30 scripts/03-exploring_cleaned_data.R
Updated summary data based on changes in the cleaned dataset

@@ -23,14 +23,14 @@ cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data-2010-2019.c
 # Generate summary statistics 
 summary_stats <- cleaned_data %>%
   summarise(
-    mean_daily_sb_cars_t = mean(daily_sb_cars_t, na.rm = TRUE),
-    median_daily_sb_cars_t = median(daily_sb_cars_t, na.rm = TRUE),
-    mean_daily_nb_cars_t = mean(daily_nb_cars_t, na.rm = TRUE),
-    median_daily_nb_cars_t = median(daily_nb_cars_t, na.rm = TRUE),
-    mean_daily_wb_cars_t = mean(daily_wb_cars_t, na.rm = TRUE),
-    median_daily_wb_cars_t = median(daily_wb_cars_t, na.rm = TRUE),
-    mean_daily_eb_cars_t = mean(daily_eb_cars_t, na.rm = TRUE),
-    median_daily_eb_cars_t = median(daily_eb_cars_t, na.rm = TRUE),
+    mean_daily_sb_cars = mean(daily_sb_cars, na.rm = TRUE),
+    median_daily_sb_cars = median(daily_sb_cars, na.rm = TRUE),
+    mean_daily_nb_cars = mean(daily_nb_cars, na.rm = TRUE),
+    median_daily_nb_cars = median(daily_nb_cars, na.rm = TRUE),
+    mean_daily_wb_cars = mean(daily_wb_cars, na.rm = TRUE),
+    median_daily_wb_cars = median(daily_wb_cars, na.rm = TRUE),
+    mean_daily_eb_cars = mean(daily_eb_cars, na.rm = TRUE),
+    median_daily_eb_cars = median(daily_eb_cars, na.rm = TRUE),
     # Buses
     mean_daily_sb_bus = mean(daily_sb_bus, na.rm = TRUE),
     median_daily_sb_bus = median(daily_sb_bus, na.rm = TRUE),
@@ -71,10 +71,10 @@ View(summary_stats)
 line_graph_car <- cleaned_data %>%
   group_by(count_date) %>%
   summarise(
-    total_sb_cars_t = sum(daily_sb_cars_t, na.rm = TRUE),
-    total_nb_cars_t = sum(daily_nb_cars_t, na.rm = TRUE),
-    total_eb_cars_t = sum(daily_eb_cars_t, na.rm = TRUE),
-    total_wb_cars_t = sum(daily_wb_cars_t, na.rm = TRUE)
[…]

### 2024-03-30 scripts/02-data_cleaning-copy.R
Created and tested model
Still needs improvements in the dataset to ensure it is sound.

@@ -1,0 +1,97 @@
+#### Preamble ####
+# Purpose: Cleans the raw traffic counts recorded into a more manageable dataset 
+# Author: Maria Mangru
+# Date: March 2024
+# Contact:maria.mangru@mail.utoronto.ca
+# License: MIT
+
+
+#### Workspace setup ####
+library(dplyr)
+library(stringr)
+
+#### Clean data ####
+
+# access raw data 
+data <- read.csv("./data/raw_data/raw-data-2010-2019.csv")
+
+# Removing less relevant columns to focus on total traffic volumes, buses, pedestrians, and cyclists
+data_simplified <- data[, c("count_id", "count_date", "location_id", "location", 
+                            "sb_cars_r","sb_cars_t","sb_cars_l","nb_cars_r", "nb_cars_t", "nb_cars_l", "wb_cars_r",
+                            "wb_cars_t","wb_cars_l", "eb_cars_r", "eb_cars_t","eb_cars_l", "sb_bus_r", "sb_bus_t", 
+                            "sb_bus_l", "nb_bus_r", "nb_bus_t", "nb_bus_l", "wb_bus_r", "wb_bus_t", "wb_bus_l", "eb_bus_r", "eb_bus_t", 
+                            "eb_bus_l", "nx_peds", "sx_peds", "ex_peds", "wx_peds", "nx_bike", "sx_bike", "ex_bike", "wx_bike", "centreline_type")]
+
+
+# Aggregating data to daily traffic volumes 
+daily_traffic_modes <- data_simplified %>%
+  group_by(count_date, location_id, location) %>%
+  summarise(daily_sb_cars = sum(sb_cars_t + sb_cars_r + sb_cars_l, na.rm = TRUE),
+            daily_nb_cars = sum(nb_cars_t + nb_cars_r + nb_cars_l, na.rm = TRUE),
+            daily_wb_cars = sum(wb_cars_t + wb_cars_r + wb_cars_l, na.rm = TRUE),
+            daily_eb_cars = sum(eb_cars_t + eb_cars_r + eb_cars_l, na.rm = TRUE),
+            daily_sb_bus = sum(sb_bus_r + sb_bus_t + sb_bus_l, na.rm = TRUE),
+            daily_nb_bus = sum(nb_bus_r + nb_bus_t + nb_bus_l, na.rm = TRUE),
[…]

### 2024-03-30 scripts/02-data_cleaning.R

@@ -8,110 +8,75 @@
 
 #### Workspace setup ####
 library(dplyr)
+library(readr)
 library(stringr)
+library(lubridate)
 
 #### Clean data ####
-colnames(data)
 
 # access raw data 
 data <- read.csv("./data/raw_data/raw-data-2010-2019.csv")
 
+# Covert dates and filter based on years 2015 to 2019
+processed_data <- data %>%
+  mutate(count_date = ymd(count_date)) %>%
+  filter(year(count_date) >= 2015, year(count_date) <= 2019)
+
+
 # Removing less relevant columns to focus on total traffic volumes, buses, pedestrians, and cyclists
-data_simplified <- data[, c("count_id", "count_date", "location_id", "location", 
+processed_data <- processed_data[, c("count_id", "count_date", "location_id", "location", 
                             "sb_cars_r","sb_cars_t","sb_cars_l","nb_cars_r", "nb_cars_t", "nb_cars_l", "wb_cars_r",
                             "wb_cars_t","wb_cars_l", "eb_cars_r", "eb_cars_t","eb_cars_l", "sb_bus_r", "sb_bus_t", 
                             "sb_bus_l", "nb_bus_r", "nb_bus_t", "nb_bus_l", "wb_bus_r", "wb_bus_t", "wb_bus_l", "eb_bus_r", "eb_bus_t", 
                             "eb_bus_l", "nx_peds", "sx_peds", "ex_peds", "wx_peds", "nx_bike", "sx_bike", "ex_bike", "wx_bike", "centreline_type")]
 
-# Remove centreline_type = 1
-#data_simplified <- data_simplified %>%
-  #filter(centreline_type != 1)
-
-
 
 # Aggregating data to daily traffic volumes 
-daily_traffic_modes <- data_simplified %>%
+daily_traffic <- processed_data %>%
   group_by(count_date, location_id, location) %>%
-  summarise(daily_sb_cars = sum(sb_cars_t + sb_cars_r + sb_cars_l, na.rm = TRUE),
-            daily_nb_cars = sum(nb_cars_t + nb_cars_r + nb_cars_l, na.rm = TRUE),
-            daily_wb_cars = sum(wb_cars_t + wb_cars_r + wb_cars_l, na.rm = TRUE),
[…]

### 2024-03-30 scripts/05-model.R

@@ -16,17 +16,65 @@ library(readr)
 #### Read data ####
 analysis_data <- read_csv("data/analysis_data/cleaned-data-2010-2019.csv")
 
-### Model data ####
-traffic_model_nb <- glm.nb(
-  formula = daily_sb_cars_t ~ daily_sb_bus + daily_wb_bus,
-  data = analysis_data,
-  link = log
-)
+## Scale the predictor variables ## 
+analysis_data$daily_bus <- scale(analysis_data$daily_bus)
+analysis_data$daily_peds <- scale(analysis_data$daily_peds)
+analysis_data$daily_bike <- scale(analysis_data$daily_bike)
 
-summary(traffic_model_nb)
+
+before_improvement <- subset(analysis_data, count_date < as.Date("2017-12-17"))
+after_improvement <- subset(analysis_data, count_date >= as.Date("2017-12-17"))
+
+
+# Check the number of observations in each subset to ensure they contain data.
+cat("Observations before improvement:", nrow(before_improvement), "\n")
+cat("Observations after improvement:", nrow(after_improvement), "\n")
+
+### Model data ####
+
+
+model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, 
+                       data = before_improvement)
+summary(model_before)
+
+
+model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, 
+                       data = before_improvement)
+summary(model_before)
 
 #### Save model ####
 saveRDS(
-  traffic_model_nb,
-  file = "model/traffic_model_nb.rds"
+  model,
+  file = "model/model.rds"
 )
+
+### Checks for Model ###
+
+library(car)
+vif(model_before) 
+vif(model_after)
+
+
+# Plotting residuals to check for patterns
+par(mfrow = c(2, 2))
+plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
+plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")
+
+
+# Checking for normality of residuals
+hist(resid(model_before), main = "Histogram of Residuals - Before")
+hist(resid(model_after), main = "Histogram of Residuals - After")
+
+
+# Checking for influential observations using Cook's distance
+cooks.distance(model_before) # For the 'before improvement' model
+cooks.distance(model_after)  # For the 'after improvement' model
+
+# Plotting Cook's distance
[…]

### 2024-03-31 scripts/02-data_cleaning-copy.R
Created tests, cleaned up code for cleaning data, exploring data and model

@@ -1,97 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw traffic counts recorded into a more manageable dataset 
-# Author: Maria Mangru
-# Date: March 2024
-# Contact:maria.mangru@mail.utoronto.ca
-# License: MIT
-
-
-#### Workspace setup ####
-library(dplyr)
-library(stringr)
-
-#### Clean data ####
-
-# access raw data 
-data <- read.csv("./data/raw_data/raw-data-2010-2019.csv")
-
-# Removing less relevant columns to focus on total traffic volumes, buses, pedestrians, and cyclists
-data_simplified <- data[, c("count_id", "count_date", "location_id", "location", 
-                            "sb_cars_r","sb_cars_t","sb_cars_l","nb_cars_r", "nb_cars_t", "nb_cars_l", "wb_cars_r",
-                            "wb_cars_t","wb_cars_l", "eb_cars_r", "eb_cars_t","eb_cars_l", "sb_bus_r", "sb_bus_t", 
-                            "sb_bus_l", "nb_bus_r", "nb_bus_t", "nb_bus_l", "wb_bus_r", "wb_bus_t", "wb_bus_l", "eb_bus_r", "eb_bus_t", 
-                            "eb_bus_l", "nx_peds", "sx_peds", "ex_peds", "wx_peds", "nx_bike", "sx_bike", "ex_bike", "wx_bike", "centreline_type")]
-
-
-# Aggregating data to daily traffic volumes 
-daily_traffic_modes <- data_simplified %>%
-  group_by(count_date, location_id, location) %>%
-  summarise(daily_sb_cars = sum(sb_cars_t + sb_cars_r + sb_cars_l, na.rm = TRUE),
-            daily_nb_cars = sum(nb_cars_t + nb_cars_r + nb_cars_l, na.rm = TRUE),
-            daily_wb_cars = sum(wb_cars_t + wb_cars_r + wb_cars_l, na.rm = TRUE),
-            daily_eb_cars = sum(eb_cars_t + eb_cars_r + eb_cars_l, na.rm = TRUE),
-            daily_sb_bus = sum(sb_bus_r + sb_bus_t + sb_bus_l, na.rm = TRUE),
-            daily_nb_bus = sum(nb_bus_r + nb_bus_t + nb_bus_l, na.rm = TRUE),
[…]

### 2024-03-31 scripts/02-data_cleaning.R

@@ -11,19 +11,20 @@ library(dplyr)
 library(readr)
 library(stringr)
 library(lubridate)
+library(here)
 
 #### Clean data ####
 
-# access raw data 
-data <- read.csv("./data/raw_data/raw-data-2010-2019.csv")
+# Reading the raw data
+data <- read.csv(here("data", "raw_data", "raw-data-2010-2019.csv"))
 
-# Covert dates and filter based on years 2015 to 2019
+# Converting dates and filtering data from 2014 to 2019
 processed_data <- data %>%
   mutate(count_date = ymd(count_date)) %>%
-  filter(year(count_date) >= 2015, year(count_date) <= 2019)
+  filter(year(count_date) >= 2014, year(count_date) <= 2019)
 
 
-# Removing less relevant columns to focus on total traffic volumes, buses, pedestrians, and cyclists
+# Removing less relevant columns to focus on total traffic volumes
 processed_data <- processed_data[, c("count_id", "count_date", "location_id", "location", 
                             "sb_cars_r","sb_cars_t","sb_cars_l","nb_cars_r", "nb_cars_t", "nb_cars_l", "wb_cars_r",
                             "wb_cars_t","wb_cars_l", "eb_cars_r", "eb_cars_t","eb_cars_l", "sb_bus_r", "sb_bus_t", 
@@ -31,7 +32,7 @@ processed_data <- processed_data[, c("count_id", "count_date", "location_id", "l
                             "eb_bus_l", "nx_peds", "sx_peds", "ex_peds", "wx_peds", "nx_bike", "sx_bike", "ex_bike", "wx_bike", "centreline_type")]
 
 
-# Aggregating data to daily traffic volumes 
+# Aggregating data to calculate daily traffic volumes
 daily_traffic <- processed_data %>%
   group_by(count_date, location_id, location) %>%
   summarise(
@@ -49,39 +50,34 @@ daily_traffic <- processed_data %>%
   ungroup()
 
 
-# Define streets to include and exclude 
-# List of streets to keep
-keep_streets <- c("SHEPPARD AVE W", "CHESSWOOD DR", "KEELE ST", "FINCH AVE W", 
-                  "TANGIERS RD", "YORK BLVD", "IAN MACDONALD BLVD", "SENTINEL RD", 
[…]

### 2024-03-31 scripts/03-exploring_cleaned_data.R

@@ -7,155 +7,94 @@
 
 
 #### Workspace setup ####
+# Load necessary libraries
 library(dplyr)
 library(ggplot2)
 library(tidyr)
 library(readr)
 library(here)
 
-#### Creating Summary Statistics and Visualizations ####
+### Creating Summary Statistics and Visualizations ###
 
 # Load the cleaned data
 cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data-2010-2019.csv"))
 
-## Summary Statistics ###
+# Summary Statistics
 
-# Generate summary statistics 
+# Generate and view summary statistics for traffic data
 summary_stats <- cleaned_data %>%
   summarise(
-    mean_daily_sb_cars = mean(daily_sb_cars, na.rm = TRUE),
-    median_daily_sb_cars = median(daily_sb_cars, na.rm = TRUE),
-    mean_daily_nb_cars = mean(daily_nb_cars, na.rm = TRUE),
-    median_daily_nb_cars = median(daily_nb_cars, na.rm = TRUE),
-    mean_daily_wb_cars = mean(daily_wb_cars, na.rm = TRUE),
-    median_daily_wb_cars = median(daily_wb_cars, na.rm = TRUE),
-    mean_daily_eb_cars = mean(daily_eb_cars, na.rm = TRUE),
-    median_daily_eb_cars = median(daily_eb_cars, na.rm = TRUE),
-    # Buses
-    mean_daily_sb_bus = mean(daily_sb_bus, na.rm = TRUE),
-    median_daily_sb_bus = median(daily_sb_bus, na.rm = TRUE),
-    mean_daily_nb_bus = mean(daily_nb_bus, na.rm = TRUE),
-    median_daily_nb_bus = median(daily_nb_bus, na.rm = TRUE),
-    mean_daily_wb_bus = mean(daily_wb_bus, na.rm = TRUE),
-    median_daily_wb_bus = median(daily_wb_bus, na.rm = TRUE),
-    mean_daily_eb_bus = mean(daily_eb_bus, na.rm = TRUE),
-    median_daily_eb_bus = median(daily_eb_bus, na.rm = TRUE),
-    # Pedestrians
-    mean_daily_nx_peds = mean(daily_nx_peds, na.rm = TRUE),
-    median_daily_nx_peds = median(daily_nx_peds, na.rm = TRUE),
-    mean_daily_sx_peds = mean(daily_sx_peds, na.rm = TRUE),
[…]

### 2024-03-31 scripts/04-test_data.R

@@ -20,70 +20,143 @@ library(here)
 ### Tests for simulated data set ###
 source(here("scripts", "00-simulate_data.R"))
 
-
-## 1: Check that simulated data set is not empty
+# Check that simulated data set is not empty
 test_that("Data frame is not empty", {
   expect_false(nrow(simulated_traffic_volumes) == 0)
 })
 
-## 2: Check for missing values in columns
+# Check for missing values in critical columns
 test_that("There are no missing values in critical columns", {
-  critical_cols <- c("count_date", "location_id", "location", "lng", "lat", "time_start", "time_end")
+  critical_cols <- c("count_date", "location_id", "location", "lng", "lat", 
+                     "time_start", "time_end")
   for (col in critical_cols) {
     expect_true(all(!is.na(simulated_traffic_volumes[[col]])))
   }
 })
 
-## 3: Check longitude and latitude values to ensure they're within Toronto's bounds
+# Check longitude and latitude values within Toronto's bounds
 test_that("Longitude and latitude are within expected range", {
-  expect_true(all(simulated_traffic_volumes$lng >= -79.63926 & simulated_traffic_volumes$lng <= -79.12206))
-  expect_true(all(simulated_traffic_volumes$lat >= 43.59052 & simulated_traffic_volumes$lat <= 43.85545))
+  expect_true(all(simulated_traffic_volumes$lng >= -79.63926 & 
+                    simulated_traffic_volumes$lng <= -79.12206))
+  expect_true(all(simulated_traffic_volumes$lat >= 43.59052 & 
+                    simulated_traffic_volumes$lat <= 43.85545))
 })
 
-## 4: Check data range to ensure they're reasonable 
+# Check count dates are within the expected range
 test_that("Count dates are within the expected range", {
   expect_true(min(simulated_traffic_volumes$count_date) >= as.Date('2010-01-01'))
   expect_true(max(simulated_traffic_volumes$count_date) <= as.Date('2019-12-31'))
 })
 
-## 5: Check that traffic counts are reasonable
+# Check that traffic counts are within reasonable limits
 test_that("Traffic counts are within reasonable limits", {
[…]

### 2024-03-31 scripts/05-model.R

@@ -9,72 +9,68 @@
 library(MASS) 
 library(dplyr)
 library(readr)
+library(car)
 
 
-## NEEDS A LOT OF WORK
+# Read and prepare data
+analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data-2010-2019.csv"))
 
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/cleaned-data-2010-2019.csv")
+# Scaling predictor variables for analysis
+analysis_data <- analysis_data %>%
+  mutate(
+    daily_bus = scale(daily_bus),
+    daily_peds = scale(daily_peds),
+    daily_bike = scale(daily_bike)
+  )
 
-## Scale the predictor variables ## 
-analysis_data$daily_bus <- scale(analysis_data$daily_bus)
-analysis_data$daily_peds <- scale(analysis_data$daily_peds)
-analysis_data$daily_bike <- scale(analysis_data$daily_bike)
 
-
+# Splitting the dataset into before and after a specific date for comparison
 before_improvement <- subset(analysis_data, count_date < as.Date("2017-12-17"))
 after_improvement <- subset(analysis_data, count_date >= as.Date("2017-12-17"))
 
 
-# Check the number of observations in each subset to ensure they contain data.
+# Verifying the data split
 cat("Observations before improvement:", nrow(before_improvement), "\n")
 cat("Observations after improvement:", nrow(after_improvement), "\n")
 
+
 ### Model data ####
 
-
-model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, 
-                       data = before_improvement)
+# Building negative binomial regression models for before and after improvement
+model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, data = before_improvement)
 summary(model_before)
 
+model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, data = after_improvement)
+summary(model_after)
 
-model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, 
-                       data = before_improvement)
-summary(model_before)
 
 #### Save model ####
-saveRDS(
-  model,
-  file = "model/model.rds"
-)
+models_dir <- here("models")
+saveRDS(model_before, file = file.path(models_dir, "model_before.rds"))
+saveRDS(model_after, file = file.path(models_dir, "model_after.rds"))
+
+
 
 ### Checks for Model ###
 
-library(car)
+# Checking for multicollinearity using VIF
 vif(model_before) 
[…]

### 2024-04-01 paper/paper.qmd
Added more data to the dataset
Added more rows of data to increase the data size the model is sampled on

@@
 date-format: long
-abstract: "In this study, I have embarked on an analysis of a decade's worth of traffic volume data from critical intersections in Toronto, scrutinizing the effects of public transit developments on urban traffic congestion from 2010 to 2019.
-[INSERT FINDINGS HERE] The significance of this research lies in its potential to inform and shape urban infrastructure strategies, offering evidence on how transit policies can alleviate traffic pressures in metropolitian areas."
+abstract: "This study investigates the relationship between enhancements in public transit and traffic volumes in metropolitan cities.
+In particular, this study focuses on the effects caused by the Spadina Subway Extension in Toronto.
+Data from 2014 - 2019 reveals that both before and after the Subway Extension in 2017, an increase in bus and pedestrian traffic was also related to an increase in car traffic."
 format: pdf

### 2024-04-01 scripts/02-data_cleaning.R

@@ -16,25 +16,48 @@ library(here)
 #### Clean data ####
 
 # Reading the raw data
-data <- read.csv(here("data", "raw_data", "raw-data-2010-2019.csv"))
+data_2010_2019 <- read.csv(here("data", "raw_data", "raw-data-2010-2019.csv"))
+data_2020_2029 <- read.csv(here("data", "raw_data", "raw-data-2020-2029.csv"))
 
-# Converting dates and filtering data from 2014 to 2019
-processed_data <- data %>%
-  mutate(count_date = ymd(count_date)) %>%
-  filter(year(count_date) >= 2014, year(count_date) <= 2019)
+# Combine datasets 
+combined_data <- rbind(data_2010_2019, data_2020_2029)
+
+
+# Check if dataset was properly combined. Code commented out after check completed.
+# head(combined_data)
+# tail(combined_data)
+# str(combined_data)
+
+
+# Converting dates and filtering data
+filtered_data <- combined_data %>%
+  filter((count_date >= as.Date("2011-12-31") & count_date <= as.Date("2014-12-31")) |
+           (count_date >= as.Date("2018-11-01") & count_date <= as.Date("2020-02-29")))
+
+filtered_data$count_date <- as.Date(filtered_data$count_date)
+
+# Creating time_trend variable
+
+# Determine the base date for the time trend 
+base_date <- min(filtered_data$count_date, na.rm = TRUE)
+
+# Adding the time_trend variable which counts days since the base date 
+filtered_data <- filtered_data %>%
+  mutate(time_trend = as.numeric(count_date - base_date))
 
 
 # Removing less relevant columns to focus on total traffic volumes
-processed_data <- processed_data[, c("count_id", "count_date", "location_id", "location", 
+processed_data <- filtered_data[, c("count_id", "count_date", "location", 
                             "sb_cars_r","sb_cars_t","sb_cars_l","nb_cars_r", "nb_cars_t", "nb_cars_l", "wb_cars_r",
                             "wb_cars_t","wb_cars_l", "eb_cars_r", "eb_cars_t","eb_cars_l", "sb_bus_r", "sb_bus_t", 
-                            "sb_bus_l", "nb_bus_r", "nb_bus_t", "nb_bus_l", "wb_bus_r", "wb_bus_t", "wb_bus_l", "eb_bus_r", "eb_bus_t", 
[…]

### 2024-04-01 scripts/03-exploring_cleaned_data.R

@@ -17,7 +17,7 @@ library(here)
 ### Creating Summary Statistics and Visualizations ###
 
 # Load the cleaned data
-cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data-2010-2019.csv"))
+cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
 
 # Summary Statistics
 
@@ -98,3 +98,23 @@ line_graph_cyclists
 
 # Save cyclist traffic plot
 ggsave("./other/outputs/cyclist_traffic_over_time.png", plot = line_graph_cyclists, width = 12, height = 6, units = "in")
+
+
+
+# Visualizing car traffic with respect to the time trend
+time_trend_graph_car <- cleaned_data %>%
+  ggplot(aes(x = time_trend, y = daily_cars)) +
+  geom_point(aes(color = daily_cars), alpha = 0.5) + 
+  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
+  scale_color_gradient(low = "blue", high = "red") +
+  labs(title = "Car Traffic Volume Over Time Trend",
+       x = "Days Since Start",
+       y = "Daily Car Traffic",
+       color = "Daily Traffic") +
+  theme_minimal()
+
+# Display car traffic plot with time trend
+time_trend_graph_car
+
+# Save the car traffic plot with time trend
+ggsave(here("other", "outputs", "car_traffic_time_trend.png"), plot = time_trend_graph_car, width = 12, height = 6, units = "in")

### 2024-04-01 scripts/04-test_data.R

@@ -78,24 +78,10 @@ test_that("Expected columns exist after data cleaning", {
 
 # Check that dates are within the filtered range of 2014 to 2019
 test_that("Dates are within the filtered range of 2014 to 2019", {
-  expect_true(all(year(daily_traffic$count_date) >= 2014 & 
-                    year(daily_traffic$count_date) <= 2019))
+  expect_true(all(year(daily_traffic$count_date) >= 2011 & 
+                    year(daily_traffic$count_date) <= 2020))
 })
 
-# Check that filtered data does not include excluded streets
-test_that("Filtered data does not include excluded streets", {
-  excluded_streets <- c("BREMNER BLVD", "FORT YORK BLVD", "SPADINA AVE", 
-                        "DUNDAS ST", "SIGNET DR", "CARPENTER RD", 
-                        "GIHON SPRING DR", "BLACK CREEK BLVD", "HARLOCK BLVD", 
-                        "ST CLAIR AVE W AT KEELE ST & WESTON RD", 
-                        "DUFFERIN ST AT STEELES AVE W", 
-                        "BLOOR ST W AT KEELE ST & PARKSIDE DR", 
-                        "BLACK CREEK DR S TCS AT JANE ST", 
-                        "HULLMAR DR AT JANE ST")
-  for (street in excluded_streets) {
-    expect_true(all(!str_detect(daily_traffic$location, regex(street, ignore_case = TRUE))))
-  }
-})
 
 
 #### Tests for exploring cleaned data ####

### 2024-04-01 scripts/05-model.R

@@ -13,7 +13,7 @@ library(car)
 
 
 # Read and prepare data
-analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data-2010-2019.csv"))
+analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
 
 # Scaling predictor variables for analysis
 analysis_data <- analysis_data %>%
@@ -25,8 +25,13 @@ analysis_data <- analysis_data %>%
 
 
 # Splitting the dataset into before and after a specific date for comparison
-before_improvement <- subset(analysis_data, count_date < as.Date("2017-12-17"))
-after_improvement <- subset(analysis_data, count_date >= as.Date("2017-12-17"))
+start_before <- as.Date("2011-12-31")
+end_before <- as.Date("2014-12-31")
+start_after <- as.Date("2018-11-01")
+end_after <- as.Date("2020-02-29")
+
+before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
+after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)
 
 
 # Verifying the data split
@@ -37,10 +42,10 @@ cat("Observations after improvement:", nrow(after_improvement), "\n")
 ### Model data ####
 
 # Building negative binomial regression models for before and after improvement
-model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, data = before_improvement)
+model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
 summary(model_before)
 
-model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike, data = after_improvement)
+model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)
 summary(model_after)
 
 

### 2024-04-01 paper/paper.qmd
Began writing up paper. Completed abstract, intro, data and parts of results

@@
 ---
-title: "An Investigation Into The Impact of Transit Improvements on Toronto Traffic Volumes 2010 - 2019"
+title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2014 - 2020"
 subtitle: "My subtitle if needed"
@@
 date-format: long
-abstract: "This study investigates the relationship between enhancements in public transit and traffic volumes in metropolitan cities.
-In particular, this study focuses on the effects caused by the Spadina Subway Extension in Toronto.
-Data from 2014 - 2019 reveals that both before and after the Subway Extension in 2017, an increase in bus and pedestrian traffic was also related to an increase in car traffic."
+abstract: "This study investigates the impact of enhanced public transit infrastucture on car congestion.
+It specifically focuses on the reintroduction of the Route 508 Lake Shore streetcar between 2014 and 2020.
+The data, sourced from the City of Toronto's Transportation Services Division, includes detailed traffic volumes for various modes at intersections across the city.
+The method of analysis includes comparing the negative binomial regression models of daily car traffic before and after transit upgrades.
+The results highlighted a significant reduction in the relationship between bus traffic and car congestion.
+Additionally, it was observed that car traffic decreased over time.
+These findings highlight the usefulness of public transit improvements in reducing urban congestion."
 format: pdf
@@
 
-```{r}
+```{r, echo=FALSE, message=FALSE}
 #| include: false
@@
 library(here)
+library(knitr)
 ```
@@
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Traffic congestion represents a significant challenge for urban cities.
+It is a non-productive activity which negatively affects economic efficiency, environmental sustainability and overall quality of life.
+To address this challenge, many cities like Toronto have undergone improvements to public transit infrastructure with the aim of reducing car dependency and mitigate traffic congestion.
+However, it is still highly debatable whether or not such improvements lead to a decrease in car congestion.
 
-The remainder of this paper is structured as follows. @sec-data....
+This study specifically investigates the 508 Lake Shore Streetcar route in Toronto.
+It utilizes two datasets provided by the City of Toronto's Transportation Services Division.
+These datasets capture traffic volumes for various modes at city intersections.
+To analyze this data, @citeR was used along with several R packages.
+Through the use of a negative binomial regression, we can assess daily car traffic relative to bus traffic, pedestrian, bicycle flows and a time trend.
+Models were created to investigate traffic patterns before and after public transit improvements.
+
+The estimand of this study is quantitatively evaluated through the change in the relationship between daily bus traffic and daily car traffic, as well as the evolution of car traffic trends over time.
+The findings from this study indicates a significant reduction in the relationship between bus traffic and car congestion, following the transit improvements.
+Additionally, a reversal from an increasing to a decreasing trend in car traffic over time was noted.
+
+This research begins with an in-depth exploration of the dataset and the analytical methods employed, highlighting the use of @citeR for data analysis.
+It then goes on to explain the results of the regression model, comparing and contrasting traffic patterns prior to and following the installation of transportation upgrades.
+The discussion interprets these findings within the broader framework of managing urban congestion and the importance of public transit.
+Through this approach, I hope to illustrate the important role of public transit improvements in combating urban congestion, as well as its significance for promoting sustainable urban mobility and imporving the standard of living in urban areas.
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+## Data Source
 
-```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
-#| echo: false
+This research relies on traffic volume data sets from the City of Toronto's Transportation Services Division.
+It accessible through the Toronto OpenData portal and is titled "Traffic Volumes at Intersections for All Modes" which is free and accessible for public use.
+The information collected in the dataset are of two main types:
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
-  theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
-```
+1.  Automatic Traffic Recorder Counts (ATR): These are segment-level volumes which capture the total number of vehicles, cyclists, or pedestrians moving in a specific direction on a street.
 
-Talk more about it.
+2.  Turning Movement Counts (TMCs): These detail the volume observed at each leg of an intersection, including the turning movement by mode (car, truck, bus, pedestrian, cyclist, other).
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+The data set covers various intersections across Toronto, providing a detailed view of the city's traffic dynamics.
+It includes data spanning 2010-2019 and 2020-2024 which was combined into one data set.
+The data set is segmented by direction of approach, turning movement, and mode, in 15-minute intervals.
+This level of detail allows for a thorough investigation of traffic patterns, which is especially important for analyzing how changes to public transport affect traffic congestion on the 508 Lake Shore route.
 
-```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
-#| warning: false
-#| message: false
+## Variables of Interst
 
-analysis_data <- read_csv(here::here("data/analysis_data/cleaned-data-2010-2019.csv"))
+In addition to a temporal trend variables, specific variables were chosen for this analysis, with an emphasis on the dynamics of car traffic in relation to bus, pedestrian, anc bicycle traffic volumes:
+
+-   Daily Car Traffic (`daily_cars`): Total volume of car traffic recorded at selected intersections along the 508 Route.
+-   Daily Bus Traffic (`daily_bus`), Pedestrian Traffic (`daily_peds`) and Bicycle Traffic (`daily_bike`): Bus, pedestrian and bike traffic volumes used to indicate the presence and frequency of non personal car usage.
+-   Time Trend (time_trend): A constructed variable to analyze traffic pattern changes over time.
+
+Using @citeR, the dataset was processed by using tools like `ggplot2` for visualization, `dplyr` for data manipulation, and `lubridate` for date management.
+These packages made it easier to clean, aggregate, and analyze traffic volumes, which helped to provide a more complex picture of traffic patterns and flow.
+
+```{r, echo=FALSE, message=FALSE, warning=FALSE}
+# Load the cleaned data
+cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
+
+# Generate and view summary statistics for traffic data
+summary_stats <- cleaned_data %>%
+  summarise(
+    mean_daily_cars = mean(daily_cars, na.rm = TRUE),
+    median_daily_cars = median(daily_cars, na.rm = TRUE),
+    mean_daily_bus = mean(daily_bus, na.rm = TRUE),
+    median_daily_bus = median(daily_bus, na.rm = TRUE),
+    mean_daily_peds = mean(daily_peds, na.rm = TRUE),
+    median_daily_peds = median(daily_peds, na.rm = TRUE),
+    mean_daily_bike = mean(daily_bike, na.rm = TRUE),
+    median_daily_bike = median(daily_bike, na.rm = TRUE)
+  )
+
+# Display the summary statistics using knitr::kable for better formatting
+knitr::kable(summary_stats, format = "html", caption = "Summary Statistics for Daily Traffic")
 
@@
 
-Talk way more about it.
+## Data Cleaning and Preparation
+
+The data set was cleaned and prepared, with yearly data files combined, relevant date periods filtered, and traffic volumes aggregated to get daily totals for every mode of transportation.
+This preprocessing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.
+
+## Summary Statistics and Graphical Analysis
 
@@
 # Results
-
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
 

### 2024-04-01 scripts/04-test_data.R

@@ -69,7 +69,8 @@ test_that("Cleaned data frame is not empty", {
 
 # Check for the existence of expected columns after cleaning
 test_that("Expected columns exist after data cleaning", {
-  expected_cols <- c("count_date", "location_id", "location", 
+  expected_cols <- c("count_date", "location", "lng", "lat", 
+                     "count_id", "time_trend", 
                      "daily_cars", "daily_bus", "daily_peds", "daily_bike")
   for (col in expected_cols) {
     expect_true(col %in% names(daily_traffic))

### 2024-04-02 paper/paper.qmd
Included data section, results and graphs
Needs fixing to expand on ideas.

@@
 title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2014 - 2020"
-subtitle: "My subtitle if needed"
+subtitle: "Examining Traffic Patterns Before and After Public Transit Improvements"
 author: 
   - Maria Mangru
-thanks: "Code and data are available at: https://github.com/MariaMangru/An-Investigation-Into-The-Impact-of-Transit-Improvements-on-Toronto-Traffic-Volumes-2010-2019."
+thanks: "Code and data are available at: https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline."
 date: today
@@
 library(knitr)
+library(stringr)
 ```
@@
 
+The data set was cleaned and prepared, with yearly data files combined, relevant date periods filtered, and traffic volumes aggregated to get daily totals for every mode of transportation.
+This preprocessing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.
+
 ## Variables of Interst
@@
 
-```{r, echo=FALSE, message=FALSE, warning=FALSE}
+```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
 # Load the cleaned data
@@
 
-# Generate and view summary statistics for traffic data
+# Generate summary statistics for traffic data
 summary_stats <- cleaned_data %>%
   summarise(
-    mean_daily_cars = mean(daily_cars, na.rm = TRUE),
-    median_daily_cars = median(daily_cars, na.rm = TRUE),
-    mean_daily_bus = mean(daily_bus, na.rm = TRUE),
-    median_daily_bus = median(daily_bus, na.rm = TRUE),
-    mean_daily_peds = mean(daily_peds, na.rm = TRUE),
-    median_daily_peds = median(daily_peds, na.rm = TRUE),
-    mean_daily_bike = mean(daily_bike, na.rm = TRUE),
-    median_daily_bike = median(daily_bike, na.rm = TRUE)
-  )
+    Mean_Cars = mean(daily_cars, na.rm = TRUE),
+    Median_Cars = median(daily_cars, na.rm = TRUE),
+    Mean_Buses = mean(daily_bus, na.rm = TRUE),
+    Median_Buses = median(daily_bus, na.rm = TRUE),
+    Mean_Pedestrians = mean(daily_peds, na.rm = TRUE),
+    Median_Pedestrians = median(daily_peds, na.rm = TRUE),
+    Mean_Bicycles = mean(daily_bike, na.rm = TRUE),
+    Median_Bicycles = median(daily_bike, na.rm = TRUE)
+  ) %>% 
+  pivot_longer(
+    cols = everything(),
+    names_to = "Category",
+    values_to = "Value"
+  ) %>%
+  separate(Category, into = c("Statistic", "Mode"), sep = "_") %>%
+  pivot_wider(
+    names_from = Statistic,
+    values_from = Value
+  ) %>%
+  mutate(Mode = str_replace(Mode, "Cars", "Daily Car Traffic"),
+         Mode = str_replace(Mode, "Buses", "Daily Bus Traffic"),
+         Mode = str_replace(Mode, "Pedestrians", "Daily Pedestrian Traffic"),
+         Mode = str_replace(Mode, "Bicycles", "Daily Bicycle Traffic")) %>%
+  select(`Transport Mode` = Mode, Mean, Median)
 
-# Display the summary statistics using knitr::kable for better formatting
-knitr::kable(summary_stats, format = "html", caption = "Summary Statistics for Daily Traffic")
+# Display summary statistics
+knitr::kable(summary_stats, format = "html", 
+             col.names = c("Transport Mode", "Mean", "Median"),
+             caption = "Summary statistics table for daily traffic volume",
+             table.attr = 'style="width:100%;"')
 
@@
 
-## Data Cleaning and Preparation
+The summary table above provides a quantitative comparison across transportation modes by condensing the traffic data for each mode into average and median statistics.
+The data reveals that car traffic surpasses other modes in volume, with the mean and median values for daily car traffic being the highest.
 
-The data set was cleaned and prepared, with yearly data files combined, relevant date periods filtered, and traffic volumes aggregated to get daily totals for every mode of transportation.
-This preprocessing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.
+```{r line-graph-cars, fig.cap="Daily Car Traffic Over Time.
+This graph displays the trend in number of cars within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
 
-## Summary Statistics and Graphical Analysis
+# Visualizing daily car traffic over time
+line_graph_car <- cleaned_data %>%
+  ggplot(aes(x = count_date, y = daily_cars)) +
+  geom_smooth(se = FALSE) +  
+  labs(title = "Daily Cars Traffic Over Time",
+       x = "Date",
+       y = "Total Cars Traffic") +
+  theme_minimal()
+
+# Display car traffic plot
+line_graph_car
+
+```
+
+The daily variation in car traffic from 2011 to 2020 is shown in Figure 1.
+The traffic volume is depicted on a smooth line graph that first increases and peaks about 2016, after which it sharply declines.
+The fall that follows may point to the benefits of the Route 508 Lake Shore streetcar improvements on easing traffic congestion.
+The peak denotes a period of increased traffic.
+
+```{r line-graph-bus, fig.cap="Daily Bus Traffic Over Time.
+This graph displays the trend in number of buses within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+
+# Visualizing daily bus traffic over time
+line_graph_buses <- cleaned_data %>%
+  ggplot(aes(x = count_date, y = daily_bus)) +
+  geom_smooth(se = FALSE) +
+  labs(title = "Daily Bus Traffic Over Time",
+       x = "Date",
+       y = "Total Bus Traffic") +
+  theme_minimal()
+
+# Display bus traffic plot
+line_graph_buses
+
+```
+
+Figure 2 depicts the daily patterns in bus traffic, which follow a similar pattern to that of car traffic.
+It is a smooth line graph that rises to a high around in 2016, then declines.
+This pattern might be an indication of how bus utilization has changed, which could be related to the streetcar route's service modifications and other improvements to public transportation.
+
+```{r line-graph-pedestrian, fig.cap="Daily Pedestrian Traffic Over Time.
+This graph displays the trend in number of pedestrian within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+
+# Visualizing daily pedestrian traffic over time
+line_graph_pedestrians <- cleaned_data %>%
+  ggplot(aes(x = count_date, y = daily_peds)) +
+  geom_smooth(se = FALSE) +
+  labs(title = "Daily Pedestrian Traffic Over Time",
+       x = "Date",
+       y = "Total Pedestrian Traffic") +
+  theme_minimal()
+
+# Display pedestrian traffic plot
+line_graph_pedestrians
+
+```
+
+Variations in pedestrian traffic over is shown in Figure 3's line graph.
+The middle of the time period has a noteworthy decrease, which is followed by a rise and another decline.
+This could be attributed to a number of things, such as changes in walkability, urban development, and perhaps the availability of better public transportation alternatives that encourage walking as a mode of transportation.
+
+```{r line-graph-bike, fig.cap="Daily Bike Traffic Over Time.
+This graph displays the trend in number of bikes within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+
+# Visualizing daily bike traffic over time
+line_graph_cyclists <- cleaned_data %>%
+  ggplot(aes(x = count_date, y = daily_bike)) +
+  geom_smooth(se = FALSE) +
+  labs(title = "Daily Cyclist Traffic Over Time",
+       x = "Date",
+       y = "Total Cyclist Traffic") +
+  theme_minimal()
+
+# Display cyclist traffic plot
+line_graph_cyclists
+
+```
+
+The pattern of bicycle traffic over time is depicted in Figure 4, where a notable rise and fall are shown on the line graph, indicating huge variations.
+The variations may be attributed to changing public opinions on riding, the implementation of bike infrastructure, or seasonal variations in weather patterns, which may be exacerbated by changes made to the transit system.
+
+```{r line-graph-time-trend, fig.cap="Daily Car Traffic Over Time.
+This graph displays the general trend in number of cars from the beginning of data collection period.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+
+# Visualizing car traffic with respect to the time trend
+time_trend_graph_car <- cleaned_data %>%
+  ggplot(aes(x = time_trend, y = daily_cars)) +
+  geom_point(aes(color = daily_cars), alpha = 0.5) + 
+  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
+  scale_color_gradient(low = "blue", high = "red") +
+  labs(title = "Car Traffic Volume Over Time Trend",
+       x = "Days Since Start",
+       y = "Daily Car Traffic",
+       color = "Daily Traffic") +
+  theme_minimal()
+
+# Display car traffic plot with time trend
+time_trend_graph_car
+
+```
+The scatter plot illustrates the fluctuating daily car counts over time, with a trend line suggesting an initial increase in traffic followed by a decline.
 
@@
 
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
 ## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
 
@@
 ```
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
 
 ### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
 
@@
 ## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
 ## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
 

### 2024-04-02 scripts/03-exploring_cleaned_data.R

@@ -24,15 +24,30 @@ cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
 # Generate and view summary statistics for traffic data
 summary_stats <- cleaned_data %>%
   summarise(
-    mean_daily_cars = mean(daily_cars, na.rm = TRUE),
-    median_daily_cars = median(daily_cars, na.rm = TRUE),
-    mean_daily_bus = mean(daily_bus, na.rm = TRUE),
-    median_daily_bus = median(daily_bus, na.rm = TRUE),
-    mean_daily_peds = mean(daily_peds, na.rm = TRUE),
-    median_daily_peds = median(daily_peds, na.rm = TRUE),
-    mean_daily_bike = mean(daily_bike, na.rm = TRUE),
-    median_daily_bike = median(daily_bike, na.rm = TRUE)
-  )
+    Mean_Cars = mean(daily_cars, na.rm = TRUE),
+    Median_Cars = median(daily_cars, na.rm = TRUE),
+    Mean_Buses = mean(daily_bus, na.rm = TRUE),
+    Median_Buses = median(daily_bus, na.rm = TRUE),
+    Mean_Pedestrians = mean(daily_peds, na.rm = TRUE),
+    Median_Pedestrians = median(daily_peds, na.rm = TRUE),
+    Mean_Bicycles = mean(daily_bike, na.rm = TRUE),
+    Median_Bicycles = median(daily_bike, na.rm = TRUE)
+  ) %>% 
+  pivot_longer(
+    cols = everything(),
+    names_to = "Category",
+    values_to = "Value"
+  ) %>%
+  separate(Category, into = c("Statistic", "Mode"), sep = "_") %>%
+  pivot_wider(
+    names_from = Statistic,
+    values_from = Value
+  ) %>%
+  mutate(Mode = str_replace(Mode, "Cars", "Daily Cars"),
+         Mode = str_replace(Mode, "Buses", "Daily Buses"),
+         Mode = str_replace(Mode, "Pedestrians", "Daily Pedestrians"),
+         Mode = str_replace(Mode, "Bicycles", "Daily Bicycles")) %>%
+  select(`Transport Mode` = Mode, Mean, Median)
 
 # Display the summary statistics
 View(summary_stats)
@@ -118,3 +133,9 @@ time_trend_graph_car
 
 # Save the car traffic plot with time trend
[…]

### 2024-04-02 paper/paper.qmd
Added information about model to paper and updated README

@@
 ---
-title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2014 - 2020"
+title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020"
 subtitle: "Examining Traffic Patterns Before and After Public Transit Improvements"
@@
 library(stringr)
+library(broom)
+library(kableExtra)
+library(MASS) 
+library(car)
 ```
@@
 # Generate summary statistics for traffic data
-summary_stats <- cleaned_data %>%
-  summarise(
+summary_stats <- summarise(cleaned_data,
   Mean_Cars = mean(daily_cars, na.rm = TRUE),
@@
   Median_Bicycles = median(daily_bike, na.rm = TRUE)
-  ) %>% 
-  pivot_longer(
-    cols = everything(),
-    names_to = "Category",
-    values_to = "Value"
-  ) %>%
-  separate(Category, into = c("Statistic", "Mode"), sep = "_") %>%
-  pivot_wider(
-    names_from = Statistic,
-    values_from = Value
-  ) %>%
-  mutate(Mode = str_replace(Mode, "Cars", "Daily Car Traffic"),
-         Mode = str_replace(Mode, "Buses", "Daily Bus Traffic"),
-         Mode = str_replace(Mode, "Pedestrians", "Daily Pedestrian Traffic"),
-         Mode = str_replace(Mode, "Bicycles", "Daily Bicycle Traffic")) %>%
-  select(`Transport Mode` = Mode, Mean, Median)
+)
+
+# Create a data frame for the kable function
+summary_stats_df <- data.frame(
+  Transport_Mode = c("Daily Car Traffic", "Daily Bus Traffic", "Daily Pedestrian Traffic", "Daily Bicycle Traffic"),
+  Mean = c(summary_stats$Mean_Cars, summary_stats$Mean_Buses, summary_stats$Mean_Pedestrians, summary_stats$Mean_Bicycles),
+  Median = c(summary_stats$Median_Cars, summary_stats$Median_Buses, summary_stats$Median_Pedestrians, summary_stats$Median_Bicycles)
+)
 
 # Display summary statistics
-knitr::kable(summary_stats, format = "html", 
+kable(summary_stats_df, format = "html", 
       col.names = c("Transport Mode", "Mean", "Median"),
@@
       table.attr = 'style="width:100%;"') 
-
 ```
@@
 ```{r line-graph-cars, fig.cap="Daily Car Traffic Over Time.
-This graph displays the trend in number of cars within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+This graph displays the trend in number of cars within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.",fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 ```{r line-graph-bus, fig.cap="Daily Bus Traffic Over Time.
-This graph displays the trend in number of buses within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+This graph displays the trend in number of buses within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 ```{r line-graph-pedestrian, fig.cap="Daily Pedestrian Traffic Over Time.
-This graph displays the trend in number of pedestrian within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+This graph displays the trend in number of pedestrian within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 ```{r line-graph-bike, fig.cap="Daily Bike Traffic Over Time.
-This graph displays the trend in number of bikes within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+This graph displays the trend in number of bikes within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 ```{r line-graph-time-trend, fig.cap="Daily Car Traffic Over Time.
-This graph displays the general trend in number of cars from the beginning of data collection period.", fig.width=10, fig.height=6, echo=FALSE, message=FALSE, warning=FALSE}
+This graph displays the general trend in number of cars from the beginning of data collection period.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 ```
+
 The scatter plot illustrates the fluctuating daily car counts over time, with a trend line suggesting an initial increase in traffic followed by a decline.
@@
 
-## Model set-up
+This study uses negative binomial regression to address the problem of overdispersion, which occurs when observed variance is greater than the mean and is frequently seen in count data.
+The dependent variable in the model is the daily count of cars, which is assumed to be impacted by temporal variations and the counts of alternative transportation modes (such as buses, bicycles, and pedestrians).
+
+As such, we estimate the following model:
 
 ```{=tex}
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+\begin{align*} 
+\log(\text{Count\_Cars}) &= \beta_0 + \beta_1 \times \text{Bus} + \beta_2 \times \text{Pedestrians} + \beta_3 \times \text{Bike} + \text{time\_trend} \\
+\text{where} & \\
+\log(\text{Count\_Cars}) & \text{ is the natural logarithm of the count of cars,} \\
+\beta_0 & \text{ is the intercept of the model,} \\
+\beta_1, \beta_2, \beta_3 & \text{ are the coefficients for the counts of buses, pedestrians, and bicycles,} \\
+\text{time\_trend} & \text{ is the variable capturing the longitudinal changes.}
+\end{align*}
 ```
 
-### Model justification
+## Model justification
+
+There are two reasons for using negative binomial regression.
+First, since events tend to cluster together in both place and time, traffic count data usually show overdispersion.
+For such data, the negative binomial model makes sense since it allows the variance to surpass the mean.
+Second, given that traffic counts frequently deviate from a normal distribution, this model can account for their skewed nature.
+
+The analysis is divided into two temporal models in order to evaluate the effects of the Route 508 Lake Shore streetcar service upgrades: `model_before`, which is the model before the improvements, and `model_after`, which is the model after the improvements.
+This method clarifies the relationship that existed before and after the policy intervention between the independent and dependent variables.
+
+Together with a temporal trend, the `model_before` captures the baseline relationship between automobile traffic and other modes of transportation.
+As an illustration of possible changes in traffic patterns brought about by the improved streetcar service, the `model_after`, on the other hand, depicts the altered dynamics following the enhancement.
+By use of comparative analysis, these models demonstrate the effectiveness of the Route 508 enhancements in mitigating car congestion.
 
 # Results
+
+```{r mdoel summaries, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
+
+# Read and prepare data
+analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
+analysis_data$count_date <- as.Date(analysis_data$count_date)
+
+# Scaling predictor variables for analysis
+analysis_data <- analysis_data %>%
+  mutate(
+    daily_bus = scale(daily_bus, center = TRUE, scale = TRUE),
+    daily_peds = scale(daily_peds, center = TRUE, scale = TRUE),
+    daily_bike = scale(daily_bike, center = TRUE, scale = TRUE)
+  )
+
+# Splitting the dataset into before and after a specific date for comparison
+start_before <- as.Date("2011-12-31")
+end_before <- as.Date("2014-12-31")
+start_after <- as.Date("2018-11-01")
+end_after <- as.Date("2020-02-29")
+
+before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
+after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)
+
+### Model data ####
+
+analysis_data <- analysis_data %>%
+  mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))
+
+before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
+after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)
+
+# Building negative binomial regression models for before and after improvement
+model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
+model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)
+
+# Tidying the model summaries 
+tidy_model_before <- tidy(model_before)
+tidy_model_after <- tidy(model_after)
+
+# Model before improvement
+kable(tidy_model_before, caption = "Negative Binomial Regression Model - Before Improvement") %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))
+
+# Model after improvement
+kable(tidy_model_after, caption = "Negative Binomial Regression Model - After Improvement") %>%
+  kable_styling(bootstrap_options = c("striped", "hover"))
+
+```
 
@@
 ## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
 

### 2024-04-02 scripts/01-download_data.R

@@ -1,43 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from opendatatoronto
-# Author: Maria Mangru
-# Date: March 2024
-# Contact:maria.mangru@mail.utoronto.ca
-# License: MIT
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(dplyr)
-library(readr)
-
-
-#### Download data ####
-
-## FIX THIS - THE DATA DOES NOT DOWNLOAD PROPERLY 
-
-# get package
-package <- show_package("traffic-volumes-at-intersections-for-all-modes")
-package
-
-# get all resources for this package
-resources <- list_package_resources("traffic-volumes-at-intersections-for-all-modes")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
-
-# load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
-
-data
-
-
-#### Save data ####
-write_csv(data, "data/raw_data/raw-data-fix.csv") 
-
-
-
-
-
-
-

### 2024-04-02 scripts/03-exploring_cleaned_data.R

@@ -32,23 +32,7 @@ summary_stats <- cleaned_data %>%
     Median_Pedestrians = median(daily_peds, na.rm = TRUE),
     Mean_Bicycles = mean(daily_bike, na.rm = TRUE),
     Median_Bicycles = median(daily_bike, na.rm = TRUE)
-  ) %>% 
-  pivot_longer(
-    cols = everything(),
-    names_to = "Category",
-    values_to = "Value"
-  ) %>%
-  separate(Category, into = c("Statistic", "Mode"), sep = "_") %>%
-  pivot_wider(
-    names_from = Statistic,
-    values_from = Value
-  ) %>%
-  mutate(Mode = str_replace(Mode, "Cars", "Daily Cars"),
-         Mode = str_replace(Mode, "Buses", "Daily Buses"),
-         Mode = str_replace(Mode, "Pedestrians", "Daily Pedestrians"),
-         Mode = str_replace(Mode, "Bicycles", "Daily Bicycles")) %>%
-  select(`Transport Mode` = Mode, Mean, Median)
-
+  ) 
 # Display the summary statistics
 View(summary_stats)
 

### 2024-04-02 scripts/05-model.R

@@ -49,6 +49,38 @@ model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_tr
 summary(model_after)
 
 
+# Graphical representation of the model results for the 'before' improvement model
+before_plot <- ggplot(before_improvement, aes(x = time_trend, y = daily_cars)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "glm.nb", formula = y ~ x, se = FALSE) + 
+  labs(title = "Relationship Between Traffic and Time Before Improvements",
+       x = "Time Trend",
+       y = "Scaled Count of Cars") +
+  theme_minimal()
+
+# Display plot
+before_plot
+
+# Save the plot
+ggsave(here("other", "outputs", "before_plot.png"), plot = time_trend_graph_car, width = 12, height = 6, units = "in")
+
+
+# Graphical representation of the model results for the 'after' improvement model
+after_plot <- ggplot(after_improvement, aes(x = time_trend, y = daily_cars)) +
+  geom_point(alpha = 0.5) +
+  geom_smooth(method = "glm.nb", formula = y ~ x, se = FALSE) +
+  labs(title = "Relationship Between Traffic and Time After Improvements",
+       x = "Time Trend",
+       y = "Scaled Count of Cars") +
+  theme_minimal()
+
+# Display plot
+after_plot
+
+# Save the plot
+ggsave(here("other", "outputs", "after_plot.png"), plot = time_trend_graph_car, width = 12, height = 6, units = "in")
+
+
 #### Save model ####
 models_dir <- here("models")
 saveRDS(model_before, file = file.path(models_dir, "model_before.rds"))

### 2024-04-17 paper/paper.qmd
began writing discussion

@@
 date-format: long
-abstract: "This study investigates the impact of enhanced public transit infrastucture on car congestion.
-It specifically focuses on the reintroduction of the Route 508 Lake Shore streetcar between 2014 and 2020.
+abstract: "This study investigates the impact of enhanced public transit infrastructure on car congestion.
+It specifically focuses on the reintroduction of the Route 508 Lake Shore streetcar between 2011 and 2020.
 The data, sourced from the City of Toronto's Transportation Services Division, includes detailed traffic volumes for various modes at intersections across the city.
@@
 library(knitr)
+library(kableExtra)
 library(stringr)
 library(broom)
-library(kableExtra)
 library(MASS) 
@@
 The discussion interprets these findings within the broader framework of managing urban congestion and the importance of public transit.
-Through this approach, I hope to illustrate the important role of public transit improvements in combating urban congestion, as well as its significance for promoting sustainable urban mobility and imporving the standard of living in urban areas.
+Through this approach, I hope to illustrate the important role of public transit improvements in combating urban congestion, as well as its significance for promoting sustainable urban mobility and improving the standard of living in urban areas.
 
@@
 The data set was cleaned and prepared, with yearly data files combined, relevant date periods filtered, and traffic volumes aggregated to get daily totals for every mode of transportation.
-This preprocessing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.
+This pre-processing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.
 
-## Variables of Interst
+## Variables of Interest
 
-In addition to a temporal trend variables, specific variables were chosen for this analysis, with an emphasis on the dynamics of car traffic in relation to bus, pedestrian, anc bicycle traffic volumes:
+In addition to a temporal trend variables, specific variables were chosen for this analysis, with an emphasis on the dynamics of car traffic in relation to bus, pedestrian, and bicycle traffic volumes:
 
@@
 
-This study uses negative binomial regression to address the problem of overdispersion, which occurs when observed variance is greater than the mean and is frequently seen in count data.
+This study uses negative binomial regression to address the problem of over-dispersion, which occurs when observed variance is greater than the mean and is frequently seen in count data.
 The dependent variable in the model is the daily count of cars, which is assumed to be impacted by temporal variations and the counts of alternative transportation modes (such as buses, bicycles, and pedestrians).
@@
 ```
-
 ## Model justification
@@
 
-```{r mdoel summaries, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
+```{r residuals-before-model, fig.cap="Residuals of the negative binomial regression model before and after the transit improvement.
+This plot helps to visualize if there are any patterns or systematic deviations unexplained by the model.", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
+# Read and prepare data
+analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
+
+# Scaling predictor variables for analysis
+analysis_data <- analysis_data %>%
+  mutate(
+    daily_bus = scale(daily_bus),
+    daily_peds = scale(daily_peds),
+    daily_bike = scale(daily_bike)
+  )
+
+# Splitting the dataset into before and after a specific date for comparison
+start_before <- as.Date("2011-12-31")
+end_before <- as.Date("2014-12-31")
+start_after <- as.Date("2018-11-01")
+end_after <- as.Date("2020-02-29")
+
+before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
+after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)
+
+### Model data ####
+
+# Building negative binomial regression models for before and after improvement
+model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
+
+model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)
+
+# Visualizing residuals to identify patterns
+par(mfrow = c(1, 2))
+plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
+plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")
+
+# Examining the normality of residuals
+# hist(resid(model_before), main = "Histogram of Residuals - Before")
+# hist(resid(model_after), main = "Histogram of Residuals - After")
+```
+
+Figures 6 and 7 show an examination of residuals from the negative binomial regression models before and after the public transit improvement.
+For the model before improvements, the residuals plot does not indicate any apparent patterns or systemic deviations, suggesting that the model's assumptions may be adequate for the data.
+After the transit improvements, the residuals' plot again shows no obvious patterns, which suggests a random scatter.
+
+\newpage
+
+```{r model summaries, echo=FALSE, message=FALSE, warning=FALSE}
 
@@
 
-### Model data ####
-
 analysis_data <- analysis_data %>%
   mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))
-
-before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
-after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)
 
@@
 # Model before improvement
-kable(tidy_model_before, caption = "Negative Binomial Regression Model - Before Improvement") %>%
-  kable_styling(bootstrap_options = c("striped", "hover"))
+kable(tidy_model_before, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - Before Improvement") %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),
+                full_width = F) %>%
+  column_spec(1, bold = TRUE, width = "3cm") %>%
+  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
+  column_spec(3, width = "3cm")
 
-# Model after improvement
-kable(tidy_model_after, caption = "Negative Binomial Regression Model - After Improvement") %>%
-  kable_styling(bootstrap_options = c("striped", "hover"))
+kable(tidy_model_after, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - After Improvement") %>%
+  kable_styling(bootstrap_options = c("striped", "hover"),
+                full_width = F) %>%
+  column_spec(1, bold = TRUE, width = "3cm") %>%
+  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
+  column_spec(3, width = "3cm")
 
 ```
+
+Table 2 represents the traffic patterns before the improvement.
+These results produced significant coefficients for both the intercept and daily bus traffic, as well as the time trend variable.
+The intercept, representing the baseline log-count of cars when all other variables are zero, was significant with an estimate of 9.340 and a standard error of 0.072, indicating a high baseline traffic volume.
+Daily bus traffic had a positive and significant relationship with car traffic, with a coefficient of 0.261, suggesting that increases in bus traffic were associated with increases in car traffic.
+The time trend also showed a significant, albeit small, positive effect on car traffic `(β = 0.000287)`, indicating a gradual increase in car traffic over time.
+
+Pedestrian and bicycle traffic, however, were not significantly associated with car traffic before the improvements (p \> 0.05 for both).
+
+Post-improvement, the model exhibits significant changes in the dynamics of traffic patterns.
+The intercept remains significant but increases to 12.283, reflecting an overall higher baseline of car traffic possibly due to other factors.
+The coefficient for daily bus traffic notably decreases to 0.096, indicating a diminished relationship between bus and car traffic after the public transit improvements.
+
+The time trend shifts from a slight increase to a significant decrease `(β = -0.001033)`, reflecting a reversal in the car traffic trend, which now shows a significant decline over time post-improvement.
+
+Variables for pedestrian and bicycle traffic remained non-significant, suggesting that these modes of transport did not have a substantial impact on car traffic volumes in either model.
+
+The model results indicate a significant reduction in the relationship between bus traffic and car congestion following the transit improvements, as shown by the decrease in the coefficient of daily bus traffic from 0.261 to 0.096.
+This suggests that improvements in the public transit system might be effectively alleviating car congestion.
+Additionally, the significant negative time trend in the post-improvement model underscores a successful overall reduction in car traffic between 2012 to 2020.
 
@@
 
+The introduction of improvements in Toronto's Route 508 Lake Shore streetcar has demonstrated a tangible reduction in the correlation between bus traffic and car congestion.
+This aligns with global evidence from cities like Amsterdam, where extensive public transit systems have significantly minimized car usage, fostering more sustainable urban environments [Source here].
+Such systems not only alleviate congestion but also reduce environmental impact by lowering emissions and promoting cleaner air.
+This study confirms the view that public transit enhancements are critical for achieving urban sustainability goals.
+
+-   Study's results highlight that improved public transit correlates with decreased car congestion.
+-   Talk about public transit and urban sustainability.
+-   Provide examples from other countries e.g., Netherlands (maybe focus on Amsterdam).
+    -   Adding any additional countries could be helpful.
+
 ## Second discussion point
+
+[Source] Upgraded public transit systems can spur economic growth by enhancing accessibility to employment and reducing travel-related expenses.
+In cities with robust public transit, property values tend to rise, and commercial areas see increased activity, illustrating the broader economic benefits of accessible public transportation.
+[Source] This study's findings underscore the potential of public transit to drive urban development by reshaping cityscapes into more pedestrian- and transit-friendly environments, thereby attracting more investment and improving the overall quality of urban life.
+
+-   Effects of improved public transit such as economic.
+-   How does public transit influence development in urban areas.
 
@@
 
+Furthermore, improving public transit infrastructure serves as a powerful tool for enhancing social equity by providing reliable and affordable transportation options to diverse populations, including the disabled, elderly, and low-income individuals.
+[Source] highlights the critical role of public transit in ensuring that mobility and access to urban opportunities are not luxuries but basic services accessible to all.
+By enabling equal participation in urban life, enhanced public transit systems can break down barriers and foster more inclusive communities.
+
+-   Public transit can enhance accessibility for disabled communities, minors, elderly, students.
+-   Public transit as an equalizer for class. i.e., people with less wealth may not be able to afford a car so improved public transit gives them access to more spaces for a lower individual cost.
+
 ## Weaknesses and next steps
+
+While this research provides valuable insights, it acknowledges several limitations that could be addressed in future studies.
+The absence of detailed demographic and socioeconomic data limits the ability to deeply understand the differential impacts of public transit improvements across various population segments.
+Furthermore, external factors such as weather conditions and broader social changes, including those prompted by global crises like the COVID-19 pandemic, were not considered but could significantly affect public transit usage and perceptions.
+
+The study's primary weakness lies in its narrow focus on traffic volume data without incorporating broader contextual factors that might influence these volumes.
+
+Future research should expand the scope to include a comparative analysis of multiple cities with varying transit modes to identify best practices and related influences.
+It would also be beneficial to integrate demographic data to explore how different groups are uniquely affected by public transit changes.
+
+-   Could improve if there is access to population data (maybe weather patterns).
+
+-   Could benefit from a comparison with different cities and modes of transit.
+
+-   Consider COVID, dangers associated with public transit and how that can affect this research which can affect attitude to public transit.
 

### 2024-04-17 scripts/05-model.R

@@ -11,9 +11,7 @@ library(dplyr)
 library(readr)
 library(car)
 library(ggplot2)
-
-cor(before_improvement[, c("daily_bus", "daily_peds", "daily_bike", "time_trend", "population")], use = "complete.obs")
-
+library(here)
 
 
 
@@ -50,7 +48,7 @@ cat("Observations after improvement:", nrow(after_improvement), "\n")
 model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
 summary(model_before)
 
-model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend + population, data = after_improvement)
+model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)
 summary(model_after)
 
 
@@ -80,7 +78,7 @@ ggplot(after_fitted, aes(x = fitted, y = observed)) +
 after_fitted
 
 # Save the plot
-# ggsave(here("other", "outputs", "before_plot.png"), plot = time_trend_graph_car, width = 12, height = 6, units = "in")
+ggsave(here("other", "outputs", "before_plot.png"), plot = time_trend_graph_car, width = 12, height = 6, units = "in")
 
 
 
@@ -114,13 +112,4 @@ cooks.distance(model_after)
 plot(cooks.distance(model_before), type = "h", main = "Cook's Distance - Before")
 plot(cooks.distance(model_after), type = "h", main = "Cook's Distance - After")
 
-
-
-# Fit a null model
-null_model <- glm.nb(daily_cars ~ 1, data = analysis_data)
-
-# Calculate McFadden's pseudo-R2
-mcfadden_r2 <- 1 - (logLik(model_before)/logLik(null_model))
-mcfadden_r2_value <- as.numeric(mcfadden_r2)  # Extract the numeric value
-print(mcfadden_r2_value)
 

### 2024-04-18 paper/paper.qmd
expanded discussion and finalized abstract

@@
 title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020"
-subtitle: "Examining Traffic Patterns Before and After Public Transit Improvements"
+subtitle: "Examining Traffic Patterns Before and After Public Transit Access was Reinstated"
 author: 
@@
 date-format: long
-abstract: "This study investigates the impact of enhanced public transit infrastructure on car congestion.
-It specifically focuses on the reintroduction of the Route 508 Lake Shore streetcar between 2011 and 2020.
-The data, sourced from the City of Toronto's Transportation Services Division, includes detailed traffic volumes for various modes at intersections across the city.
-The method of analysis includes comparing the negative binomial regression models of daily car traffic before and after transit upgrades.
-The results highlighted a significant reduction in the relationship between bus traffic and car congestion.
-Additionally, it was observed that car traffic decreased over time.
-These findings highlight the usefulness of public transit improvements in reducing urban congestion."
+abstract: "This study investigates the impact of access to the Route 508 Lake Shore streetcar in Toronto on reducing urban car congestion from 2011 to 2020.
+By analyzing traffic volume data from the City of Toronto's Transportation Services Division and employing negative binomial regression models, we observed a notable shift in travel behavior, highlighted by a significant decrease in the correlation between bus traffic and car congestion.
+Specifically, the coefficient for daily bus traffic reduced from 0.261 to 0.096 after access was reinstated, indicating a reduction in car traffic as more public transit options became available.
+This highlights the effectiveness of enhancing public transit infrastructure not only in easing car congestion but also in contributing to broader environmental and economic benefits.
+By improving public transit, cities can potentially achieve sustainable development goals by providing equitable access to essential services and reducing environmental impacts, thereby improving quality of life for all urban residents."
 format: pdf
@@
 
-## First discussion point {#sec-first-point}
+## Reducing urban congestion and advancing sustainability through public transit
 
-The introduction of improvements in Toronto's Route 508 Lake Shore streetcar has demonstrated a tangible reduction in the correlation between bus traffic and car congestion.
-This aligns with global evidence from cities like Amsterdam, where extensive public transit systems have significantly minimized car usage, fostering more sustainable urban environments [Source here].
-Such systems not only alleviate congestion but also reduce environmental impact by lowering emissions and promoting cleaner air.
-This study confirms the view that public transit enhancements are critical for achieving urban sustainability goals.
+This study’s analysis, centered on the Route 508 Lake Shore streetcar enhancements in Toronto, demonstrates a statistically significant divergence of bus traffic from car congestion trends, suggesting that improvements to public transit infrastructure can effectively reduce car congestion in urban areas.
+Specifically a decrease in the coefficient for `daily_bus` from 0.261 to 0.096 post-improvement, indicates a notable shift in travel behaviour.
 
--   Study's results highlight that improved public transit correlates with decreased car congestion.
--   Talk about public transit and urban sustainability.
--   Provide examples from other countries e.g., Netherlands (maybe focus on Amsterdam).
-    -   Adding any additional countries could be helpful.
+These local findings align with research conducted by Verbavatz and Barthelemy `source` in which they found that urban area size and public transport density are pivotal in managing car traffic in cities, rather than urban density alone.
+The article further states that enhancing the density and accessibility of public transport is a viable path to mitigating traffic and its associated environmental impacts.
 
-## Second discussion point
+Moreover, the benefits of improved public transit does not stop at decreased car congestion.
+Insights from the American Public Transportation Association, which illustrate that public transit investments yield substantial congestion relief and economic benefits.
+Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested `source`.
+While on the individual level, households in America can save more than \$13,000 per year by using public transit and owning one less car.
 
-[Source] Upgraded public transit systems can spur economic growth by enhancing accessibility to employment and reducing travel-related expenses.
-In cities with robust public transit, property values tend to rise, and commercial areas see increased activity, illustrating the broader economic benefits of accessible public transportation.
-[Source] This study's findings underscore the potential of public transit to drive urban development by reshaping cityscapes into more pedestrian- and transit-friendly environments, thereby attracting more investment and improving the overall quality of urban life.
+These advantages of public transportation investments are in alignment with the principles of sustainable development as defined by the UN's World Commission on Environment and Development: to meet current needs without compromising the ability of future generations to meet their own needs `source`.
+Together, these strands of evidence support the case for enhancing public transit as a strategic approach to improving urban traffic conditions and achieving broader sustainability goals.
 
--   Effects of improved public transit such as economic.
--   How does public transit influence development in urban areas.
+## Good public transit provides access to essential services and promotes social equity
 
-## Third discussion point
+Public transit systems supports not only the movement of people from place to place but also acts a vital channel for economic empowerment and social integration.
+Access to jobs, education, healthcare, and cultural institutions are fundamental rights, and equitable public transit policies are crucial to ensuring that these services are available to all individuals, regardless of their socioeconomic status.
 
-Furthermore, improving public transit infrastructure serves as a powerful tool for enhancing social equity by providing reliable and affordable transportation options to diverse populations, including the disabled, elderly, and low-income individuals.
-[Source] highlights the critical role of public transit in ensuring that mobility and access to urban opportunities are not luxuries but basic services accessible to all.
-By enabling equal participation in urban life, enhanced public transit systems can break down barriers and foster more inclusive communities.
+Supporting this claim, Lexer Quamie's research (2011) `source` highlights the integral role of transportation policy in civil rights, particularly in periods of pronounced unemployment and income disparity.
+Quamie emphasizes that transportation decisions significantly affect who can access employment opportunities and essential services, positioning public transit as a pivotal tool in the fight for civil rights.
+By facilitating or hindering access to critical resources, transportation policies not only impact economic outcomes but also health and environmental conditions.
+This is evident as inadequate or unaffordable public transit disproportionately affects low-income populations, limiting their ability to reach jobs, healthcare, and thereby affecting their overall economic and health status.
+Additionally, prioritization of car travel over public transit can lead to increased emissions, affecting urban air quality and contributing to environmental degradation.
 
--   Public transit can enhance accessibility for disabled communities, minors, elderly, students.
--   Public transit as an equalizer for class. i.e., people with less wealth may not be able to afford a car so improved public transit gives them access to more spaces for a lower individual cost.
+Moreover, the significance of social equity in public transit is underscored by the stigma associated with bus ridership, particularly among racial minorities and economically marginalized groups.
+This stigma can lead to policies that inadvertently reinforce social inequities.
+Actively combating such biases, transit policymakers can ensure that public transit remains an inclusive service that aligns with the broader goals of sustainable development.
 
@@
 
+### Weaknesses
+
 While this research provides valuable insights, it acknowledges several limitations that could be addressed in future studies.
 The absence of detailed demographic and socioeconomic data limits the ability to deeply understand the differential impacts of public transit improvements across various population segments.
-Furthermore, external factors such as weather conditions and broader social changes, including those prompted by global crises like the COVID-19 pandemic, were not considered but could significantly affect public transit usage and perceptions.
+Moreover, external factors like varying weather conditions and significant social shifts such as attitudes towards public transit, have not been considered.
+These elements have the potential to influence public transit usage patterns and public perception significantly.
+One of the study's primary limitations is its narrow analytical lens, concentrated on traffic volume data, which potentially overlooks the other factors that might influence traffic patterns.
+For instance, changes in the local economy or shifts in population density.
 
-The study's primary weakness lies in its narrow focus on traffic volume data without incorporating broader contextual factors that might influence these volumes.
+### Next Steps
 
-Future research should expand the scope to include a comparative analysis of multiple cities with varying transit modes to identify best practices and related influences.
-It would also be beneficial to integrate demographic data to explore how different groups are uniquely affected by public transit changes.
-
--   Could improve if there is access to population data (maybe weather patterns).
-
--   Could benefit from a comparison with different cities and modes of transit.
-
--   Consider COVID, dangers associated with public transit and how that can affect this research which can affect attitude to public transit.
+Future research directions should broaden the analytical framework to incorporate a comparative analysis of diverse urban settings, including multiple transit modes to identify best practices and the varying impacts these have across different population segments.
+Additionally, it would be beneficial to integrate granular demographic data, enabling a deeper exploration of how public transit enhancements affect population groups defined by age, income, race, and other socioeconomic factors.
 

### 2024-04-18 paper/paper.qmd
fixed up intro and data source sections

@@
 ---
-title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020"
+title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2019"
 subtitle: "Examining Traffic Patterns Before and After Public Transit Access was Reinstated"
@@
 
-Traffic congestion represents a significant challenge for urban cities.
-It is a non-productive activity which negatively affects economic efficiency, environmental sustainability and overall quality of life.
-To address this challenge, many cities like Toronto have undergone improvements to public transit infrastructure with the aim of reducing car dependency and mitigate traffic congestion.
-However, it is still highly debatable whether or not such improvements lead to a decrease in car congestion.
+Traffic congestion represents a significant challenge for urban areas, negatively affecting economic productivity, environmental sustainability, and quality of life.
+In cities like Toronto, public transit improvements are often implemented as a strategic response to reduce reliance on personal vehicles and mitigate congestion.
+This study centers on the Route 508 Lake Shore streetcar in Toronto, examining its impact on urban traffic dynamics, particularly in reducing car congestion.
 
-This study specifically investigates the 508 Lake Shore Streetcar route in Toronto.
-It utilizes two datasets provided by the City of Toronto's Transportation Services Division.
-These datasets capture traffic volumes for various modes at city intersections.
-To analyze this data, @citeR was used along with several R packages.
-Through the use of a negative binomial regression, we can assess daily car traffic relative to bus traffic, pedestrian, bicycle flows and a time trend.
-Models were created to investigate traffic patterns before and after public transit improvements.
+The primary aim of this research is to evaluate the effect of the 508 Lake Shore streetcar enhancements on daily traffic patterns, focusing on how these changes influence the interaction between car traffic and other transportation modes such as buses, bicycles, and pedestrians.
+By employing traffic volume data collected by the City of Toronto's Transportation Services Division, we utilize negative binomial regression models to quantify the shifts in traffic patterns from 2011 to 2020.
+The estimand of this study is the change in the correlation between daily bus traffic and daily car traffic following the transit improvements.
 
-The estimand of this study is quantitatively evaluated through the change in the relationship between daily bus traffic and daily car traffic, as well as the evolution of car traffic trends over time.
-The findings from this study indicates a significant reduction in the relationship between bus traffic and car congestion, following the transit improvements.
-Additionally, a reversal from an increasing to a decreasing trend in car traffic over time was noted.
+The data source used is the Toronto OpenData portal, which provides comprehensive traffic counts across the city, making the analysis both robust and replicable.
+The data set, "Traffic Volumes at Intersections for All Modes," offers a granular view of the city's traffic flows at various intersections in the city.
 
-This research begins with an in-depth exploration of the dataset and the analytical methods employed, highlighting the use of @citeR for data analysis.
-It then goes on to explain the results of the regression model, comparing and contrasting traffic patterns prior to and following the installation of transportation upgrades.
-The discussion interprets these findings within the broader framework of managing urban congestion and the importance of public transit.
-Through this approach, I hope to illustrate the important role of public transit improvements in combating urban congestion, as well as its significance for promoting sustainable urban mobility and improving the standard of living in urban areas.
+This investigation not only sheds light on the specific case of the Route 508 Lake Shore but aims to contribute to a broader understanding of how enhancing public transit infrastructure can significantly alter urban mobility patterns.
+
+The remainder of this paper is structured as follows: @sec-data introduces the data used for analysis, detailing the sources, preparation, and specific variables considered, alongside initial descriptive statistics and visualizations.
+@sec-model presents the negative binomial regression model, justifying its use over other models and detailing the statistical methods employed to analyze the traffic volume data.
+@sec-results displays the results of the analysis, interpreting the changes in traffic patterns before and after the public transit improvements, with a focus on the correlation between different modes of transportation.
+@sec-discussion provides a discussion on the implications of the findings, considering the broader context of urban congestion management and the role of public transit in promoting sustainable urban mobility.
+This section also outlines the limitations of the current study and suggests areas for further research.
 
@@
 
-This research relies on traffic volume data sets from the City of Toronto's Transportation Services Division.
-It accessible through the Toronto OpenData portal and is titled "Traffic Volumes at Intersections for All Modes" which is free and accessible for public use.
-The information collected in the dataset are of two main types:
+This research relies on traffic volume provided by the City of Toronto's Transportation Services Division, available through the Toronto OpenData portal.
+The dataset, titled "Traffic Volumes at Intersections for All Modes," is an open-access resource under the Open Government License - Toronto, ensuring broad usability and transparency.
+The data files for the years 2010-2019 were utilized, however, the data files are uploaded daily with data ranging from 1980 to present.
+The traffic volume data are cross-referenced with the City of Toronto's Street Centreline, Intersection File, and Street Traffic Signal datasets.
 
-1.  Automatic Traffic Recorder Counts (ATR): These are segment-level volumes which capture the total number of vehicles, cyclists, or pedestrians moving in a specific direction on a street.
+The data consist of two primary types of traffic counts, which capture detailed movement patterns across the city:
 
-2.  Turning Movement Counts (TMCs): These detail the volume observed at each leg of an intersection, including the turning movement by mode (car, truck, bus, pedestrian, cyclist, other).
+1.
+Automatic Traffic Recorder (ATR) Counts: These counts measure segment-level volumes, capturing the movement of vehicles, cyclists, and pedestrians in specific directions along city streets.
+For example, ATR data might include the total number of eastbound cyclists on Queen Street.
 
-The data set covers various intersections across Toronto, providing a detailed view of the city's traffic dynamics.
-It includes data spanning 2010-2019 and 2020-2024 which was combined into one data set.
-The data set is segmented by direction of approach, turning movement, and mode, in 15-minute intervals.
-This level of detail allows for a thorough investigation of traffic patterns, which is especially important for analyzing how changes to public transport affect traffic congestion on the 508 Lake Shore route.
+2.
+Turning Movement Counts (TMCs): These counts provide granular details at intersections, recording the volume of traffic movement (including turns) by various modes such as cars, trucks, buses, and bicycles.
+Each TMC includes data from eight non-continuous hours throughout a day, reported in 15-minute intervals.
 
-The data set was cleaned and prepared, with yearly data files combined, relevant date periods filtered, and traffic volumes aggregated to get daily totals for every mode of transportation.
-This pre-processing step made sure the data was suitable for analysis, with a particular emphasis on the time frame prior to and following the 508 Lake Shore route's streetcar upgrades.
+The analysis in this paper was conducted using the R programming language @citeR, which facilitated the statistical computations and data visualization.
+Further, a suite of R packages, including several from the tidyverse collection were utilized.
+The `dplyr` package was used the transformation and summarization of traffic data.
+For importing and processing string data, `readr` and `stringr` were utilized.
+The handling of date and time variables was done using `lubridate` and data visualization was achieved with `ggplot2`, allowing us to explore and reveal traffic patterns visually.
+Additionally, `tidyr` assisted in reshaping the data to ensure it was structured appropriately for analysis.
+File path management was handled using the `here` package, which was essential for ensuring reproducibility across different computing environments.
+
+The dataset was refined to aggregate daily totals for each mode of transportation, with the data aligned and merged based on date, location, and traffic mode specifics.
 
@@
 -   Time Trend (time_trend): A constructed variable to analyze traffic pattern changes over time.
-
-Using @citeR, the dataset was processed by using tools like `ggplot2` for visualization, `dplyr` for data manipulation, and `lubridate` for date management.
-These packages made it easier to clean, aggregate, and analyze traffic volumes, which helped to provide a more complex picture of traffic patterns and flow.
 
@@
 
-# Model
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-results}
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-04-18 scripts/02-data_cleaning.R

@@ -102,3 +102,4 @@ daily_traffic <- daily_traffic %>%
 write.csv(daily_traffic, here("data", "analysis_data", "cleaned-data.csv"), row.names = FALSE)
 
   
+colnames(raw_data_2010_2019)

### 2024-04-18 scripts/03-exploring_cleaned_data.R

@@ -122,4 +122,3 @@ ggsave(here("other", "outputs", "car_traffic_time_trend.png"), plot = time_trend
 
 
 
-

### 2024-04-18 paper/paper.qmd
completed discussion and model results, updated references and included literature

@@
 ---
-title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2019"
+title: "The Route 508 Streetcar and Congestion on Toronto's Shoreline, 2011 - 2020"
 subtitle: "Examining Traffic Patterns Before and After Public Transit Access was Reinstated"
@@
 By improving public transit, cities can potentially achieve sustainable development goals by providing equitable access to essential services and reducing environmental impacts, thereby improving quality of life for all urban residents."
-format: pdf
+format: 
+  pdf:
+    documentclass: article
+    geometry: [margin=1in]
 number-sections: true
@@
 
-```{r, echo=FALSE, message=FALSE}
+```{r, include= FALSE, echo=FALSE, message=FALSE}
 #| include: false
@@
 library(car)
+library(patchwork)
+```
+
+```{r, include= FALSE, echo=FALSE, message=FALSE}
+
+#Import cleaned data 
+analysis_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
+
 ```
@@
 The estimand of this study is the change in the correlation between daily bus traffic and daily car traffic following the transit improvements.
+For the purposes of this research, the terms "improvements" and "reinstating of the streetcar" are used interchangeably.
+Both terms refer to the suite of upgrades and changes implemented along the 508 Lake Shore route, which included the restoration of streetcar services that had previously been discontinued or altered.'
 
@@
 The dataset, titled "Traffic Volumes at Intersections for All Modes," is an open-access resource under the Open Government License - Toronto, ensuring broad usability and transparency.
-The data files for the years 2010-2019 were utilized, however, the data files are uploaded daily with data ranging from 1980 to present.
+The data files for the years 2010-2019 and 2020-2029 were utilized, however, the data files are uploaded daily with data ranging from 1980 to present.
 The traffic volume data are cross-referenced with the City of Toronto's Street Centreline, Intersection File, and Street Traffic Signal datasets.
@@
 Further, a suite of R packages, including several from the tidyverse collection were utilized.
-The `dplyr` package was used the transformation and summarization of traffic data.
-For importing and processing string data, `readr` and `stringr` were utilized.
-The handling of date and time variables was done using `lubridate` and data visualization was achieved with `ggplot2`, allowing us to explore and reveal traffic patterns visually.
-Additionally, `tidyr` assisted in reshaping the data to ensure it was structured appropriately for analysis.
-File path management was handled using the `here` package, which was essential for ensuring reproducibility across different computing environments.
+The @dplyr package was used the transformation and summarization of traffic data.
+For importing and processing string data, @readr and @stringr were utilized.
+The handling of date and time variables was done using @lubridate and data visualization was achieved with @ggplot2, allowing us to explore and reveal traffic patterns visually.
+Additionally, @tidyverse assisted in reshaping the data to ensure it was structured appropriately for analysis.
+File path management was handled using the @here package, which was essential for ensuring reproducibility across different computing environments.
 
 The dataset was refined to aggregate daily totals for each mode of transportation, with the data aligned and merged based on date, location, and traffic mode specifics.
+While the dataset was merged from two data files which span from 2010 to 2019 and 2020 to 2029, this research will focus on December 2011 to December 2014 and November 2018 to February 2020.
+These years were chosen to provide a comprehensive view of traffic patterns prior to, during, and following the enhancement of the 508 Lake Shore route.
+The data collection period was intentionally concluded in February 2020 to ensure that the results were not influenced by the disruptions caused by the COVID-19 pandemic.
 
@@
 
-In addition to a temporal trend variables, specific variables were chosen for this analysis, with an emphasis on the dynamics of car traffic in relation to bus, pedestrian, and bicycle traffic volumes:
-
--   Daily Car Traffic (`daily_cars`): Total volume of car traffic recorded at selected intersections along the 508 Route.
--   Daily Bus Traffic (`daily_bus`), Pedestrian Traffic (`daily_peds`) and Bicycle Traffic (`daily_bike`): Bus, pedestrian and bike traffic volumes used to indicate the presence and frequency of non personal car usage.
--   Time Trend (time_trend): A constructed variable to analyze traffic pattern changes over time.
+In this study, the examination of traffic dynamics is conducted by analyzing interactions among various modes of transportation at various intersections.
+As such, traffic counts for each mode of transit were aggregated to daily totals to allow for better analysis.
+For example, car traffic variables such as `sb_cars_l`, `nb_cars_r`, `eb_cars_t`, among others, representing different directions and types of car movements at intersections, were combined to form the Daily Car Traffic (`daily_cars`).
+This variable captures the total volume of car traffic and serves as a primary indicator of vehicular congestion.
+Similarly, this was done to create the variable Daily Bus Traffic (`daily_bus`), which helps in evaluating the volume and effectiveness of bus transit in relation to car traffic.
+Furthermore, Daily Pedestrian Traffic (`daily_peds`) and Daily Bicycle Traffic (`daily_bike`) were created to aid in understanding the prevalence and trends of non-vehicular traffic modes.
+A Time Trend variable (`time_trend`) was constructed to analyze the longitudinal changes in traffic patterns over time, providing insights into the temporal effects of transit improvements.
 
 ```{r summary-stats-table, echo=FALSE, message=FALSE, warning=FALSE}
-# Load the cleaned data
-cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
+#| label: tbl-trafficstats
+#| tbl-cap: Summary statistics of Daily Traffic Volume by mode of transportation. This table displays the mean and median counts of cars, buses, pedestrians, and bicycles
 
 # Generate summary statistics for traffic data
-summary_stats <- summarise(cleaned_data,
+summary_stats <- summarise(analysis_data,
   Mean_Cars = mean(daily_cars, na.rm = TRUE),
@@
 
-The summary table above provides a quantitative comparison across transportation modes by condensing the traffic data for each mode into average and median statistics.
-The data reveals that car traffic surpasses other modes in volume, with the mean and median values for daily car traffic being the highest.
+The summary statistics presented in @tbl-trafficstats offers a quantitative comparison across different modes of transportation, capturing the daily traffic volumes on the 508 Lake Shore route.
+Notably, **Daily Car Traffic** significantly surpasses other modes of transit, with mean and median values at 14,169.85 and 13,291.5, respectively.
+This highlights the predominant use of cars in the area, highlighting the challenges of reducing vehicular congestion.
 
-```{r line-graph-cars, fig.cap="Daily Car Traffic Over Time.
-This graph displays the trend in number of cars within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.",fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
+In contrast, **Daily Bus Traffic** shows much lower volumes, with a mean of 223.91 and a median of 190.5, reflecting less frequent usage but indicating a critical focus area for transit improvements.
+**Daily Pedestrian Traffic** and **Daily Bicycle Traffic** also exhibit considerably lower traffic volumes, with means of 686.06 and 50.49, respectively.
+The pedestrian traffic presents a median significantly lower than its mean, at 344, suggesting a skewed distribution with potentially high traffic on certain days or specific intersections.
+Similarly, the median for bicycle traffic stands at 22, indicating that on many days, bicycle usage is quite low despite a higher average, which might reflect seasonal variations or special events influencing bike use.
 
-# Visualizing daily car traffic over time
-line_graph_car <- cleaned_data %>%
+\newpage
+
+```{r}
+#| label: fig-trafficplots
+#| fig-cap: Traffic Trends by Mode along the 508 Lakeshore Route from December 2011 to December 2014 and November 2018 to February 2020.
+#| echo: FALSE
+#| warning: FALSE
+#| message: FALSE
+
+# Base theme for all plots 
+base_theme <- theme_minimal() +
+  theme(plot.title = element_text(size = 10), 
+        axis.title = element_text(size = 8), 
+        axis.text = element_text(size = 6))
+
+# Car Traffic Plot
+line_graph_car <- analysis_data %>%
   ggplot(aes(x = count_date, y = daily_cars)) +
   geom_smooth(se = FALSE) +  
-  labs(title = "Daily Cars Traffic Over Time",
+  labs(title = "Car Traffic",
        x = "Date",
-       y = "Total Cars Traffic") +
-  theme_minimal()
+       y = "Total Cars") +
+  base_theme
 
-# Display car traffic plot
-line_graph_car
+# Bus Traffic Plot
+line_graph_buses <- analysis_data %>%
+  ggplot(aes(x = count_date, y = daily_bus)) +
+  geom_smooth(se = FALSE) +
+  labs(title = "Bus Traffic",
+       x = "Date",
+       y = "Total Buses") +
+  base_theme
+
+# Pedestrian Traffic Plot
+line_graph_pedestrians <- analysis_data %>%
+  ggplot(aes(x = count_date, y = daily_peds)) +
+  geom_smooth(se = FALSE) +
+  labs(title = "Pedestrian Traffic",
+       x = "Date",
+       y = "Total Pedestrians") +
+  base_theme
+
+# Cyclist Traffic Plot
+line_graph_cyclists <- analysis_data %>%
+  ggplot(aes(x = count_date, y = daily_bike)) +
+  geom_smooth(se = FALSE) +
+  labs(title = "Cyclist Traffic",
+       x = "Date",
+       y = "Total Cyclists") +
+  base_theme
+
+# Combine plots 
+combined_plots <- (line_graph_car | line_graph_buses) / 
+                  (line_graph_pedestrians | line_graph_cyclists) & 
+                  plot_layout(guides = 'collect')
+
+combined_plots
 
@@
 
-The daily variation in car traffic from 2011 to 2020 is shown in Figure 1.
-The traffic volume is depicted on a smooth line graph that first increases and peaks about 2016, after which it sharply declines.
-The fall that follows may point to the benefits of the Route 508 Lake Shore streetcar improvements on easing traffic congestion.
-The peak denotes a period of increased traffic.
+@fig-trafficplots presents a visual analysis of the various modes of traffic from 2011 to 2020, demonstrating their respective trends and allowing for comparative insights.
+The car traffic graph illustrates an initial uptrend in volume, reaching a peak around 2016, followed by a pronounced decline that coincides with the reintroduction of the Route 508 Lake Shore streetcar, suggesting a potential correlation with improved public transit and a reduction in car use.
+Concurrently, bus traffic mirrors this pattern to a degree, with a rise and subsequent fall post-2016, hinting at an adaptive response in public transport usage.
+The pedestrian traffic trendline depicts a mid-period dip followed by a recovery and then a decline again.
+Cyclist traffic, on the other hand, displays a waveform pattern with significant ups and downs, possibly reflecting the impact of varied factors including public attitudes towards cycling, the introduction or enhancement of cycling infrastructure, and seasonal changes that affect cycling prevalence.
 
-```{r line-graph-bus, fig.cap="Daily Bus Traffic Over Time.
-This graph displays the trend in number of buses within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
+\newpage
 
-# Visualizing daily bus traffic over time
-line_graph_buses <- cleaned_data %>%
-  ggplot(aes(x = count_date, y = daily_bus)) +
-  geom_smooth(se = FALSE) +
-  labs(title = "Daily Bus Traffic Over Time",
-       x = "Date",
-       y = "Total Bus Traffic") +
-  theme_minimal()
-
-# Display bus traffic plot
-line_graph_buses
-
-```
-
-Figure 2 depicts the daily patterns in bus traffic, which follow a similar pattern to that of car traffic.
-It is a smooth line graph that rises to a high around in 2016, then declines.
-This pattern might be an indication of how bus utilization has changed, which could be related to the streetcar route's service modifications and other improvements to public transportation.
-
-```{r line-graph-pedestrian, fig.cap="Daily Pedestrian Traffic Over Time.
-This graph displays the trend in number of pedestrian within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
-
-# Visualizing daily pedestrian traffic over time
-line_graph_pedestrians <- cleaned_data %>%
-  ggplot(aes(x = count_date, y = daily_peds)) +
-  geom_smooth(se = FALSE) +
-  labs(title = "Daily Pedestrian Traffic Over Time",
-       x = "Date",
-       y = "Total Pedestrian Traffic") +
-  theme_minimal()
-
-# Display pedestrian traffic plot
-line_graph_pedestrians
-
-```
-
-Variations in pedestrian traffic over is shown in Figure 3's line graph.
-The middle of the time period has a noteworthy decrease, which is followed by a rise and another decline.
-This could be attributed to a number of things, such as changes in walkability, urban development, and perhaps the availability of better public transportation alternatives that encourage walking as a mode of transportation.
-
-```{r line-graph-bike, fig.cap="Daily Bike Traffic Over Time.
-This graph displays the trend in number of bikes within the area of interest between December 2011 to December 2014 and November 2018 to February 2020.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
-
-# Visualizing daily bike traffic over time
-line_graph_cyclists <- cleaned_data %>%
-  ggplot(aes(x = count_date, y = daily_bike)) +
-  geom_smooth(se = FALSE) +
-  labs(title = "Daily Cyclist Traffic Over Time",
-       x = "Date",
-       y = "Total Cyclist Traffic") +
-  theme_minimal()
-
-# Display cyclist traffic plot
-line_graph_cyclists
-
-```
-
-The pattern of bicycle traffic over time is depicted in Figure 4, where a notable rise and fall are shown on the line graph, indicating huge variations.
-The variations may be attributed to changing public opinions on riding, the implementation of bike infrastructure, or seasonal variations in weather patterns, which may be exacerbated by changes made to the transit system.
-
-```{r line-graph-time-trend, fig.cap="Daily Car Traffic Over Time.
-This graph displays the general trend in number of cars from the beginning of data collection period.", fig.width=8, fig.height=4, echo=FALSE, message=FALSE, warning=FALSE}
+```{r}
+#| label: fig-cartimegraph
+#| fig-cap: Daily Car Traffic Over Time. This graph displays the general trend in number of cars from the beginning of data collection period.
+#| echo: FALSE
+#| warning: FALSE
+#| message: FALSE
 
 # Visualizing car traffic with respect to the time trend
-time_trend_graph_car <- cleaned_data %>%
+time_trend_graph_car <- analysis_data %>%
   ggplot(aes(x = time_trend, y = daily_cars)) +
@@
 
-The scatter plot illustrates the fluctuating daily car counts over time, with a trend line suggesting an initial increase in traffic followed by a decline.
+The scatter plot in @fig-cartimegraph, representing daily car traffic counts over time against a time trend, offers a granular view of the data.
+Each dot signifies the traffic volume on a single day, while the trend line captures the general trajectory over the entire data collection period.
+An initial increase in traffic volume is apparent, eventually giving way to a downward trend.
 
@@
 
-This study uses negative binomial regression to address the problem of over-dispersion, which occurs when observed variance is greater than the mean and is frequently seen in count data.
-The dependent variable in the model is the daily count of cars, which is assumed to be impacted by temporal variations and the counts of alternative transportation modes (such as buses, bicycles, and pedestrians).
+Modeling traffic dynamics poses unique challenges, particularly with count data that often exhibit over-dispersion—where the variance exceeds the mean—making conventional regression models like Poisson unsuitable.
+To address this, this study employs a negative binomial regression model in order to handle over-dispersion and more aligned with the distribution characteristics of traffic counts.
+
+We define our model with the daily car traffic count as the dependent variable, hypothesizing its dependence on several factors: temporal trends and daily counts of alternative transport modes—buses, pedestrians, and bicycles.
 
@@
 
-There are two reasons for using negative binomial regression.
-First, since events tend to cluster together in both place and time, traffic count data usually show overdispersion.
-For such data, the negative binomial model makes sense since it allows the variance to surpass the mean.
-Second, given that traffic counts frequently deviate from a normal distribution, this model can account for their skewed nature.
+The rationale for selecting the negative binomial model is based on two primary areas.
+Firstly, traffic data are prone to clustering in both spatial and temporal dimensions, leading to overdispersion.
+Secondly, traffic count data typically skew away from a normal distribution, and the negative binomial model is equipped to address this skewness.
 
-The analysis is divided into two temporal models in order to evaluate the effects of the Route 508 Lake Shore streetcar service upgrades: `model_before`, which is the model before the improvements, and `model_after`, which is the model after the improvements.
-This method clarifies the relationship that existed before and after the policy intervention between the independent and dependent variables.
+To capture the effects of the Route 508 Lake Shore streetcar service reinstatement, the analysis dives into two temporal segments: pre-improvement (`model_before`) and post-improvement (`model_after`).
+This segmentation enables a nuanced understanding of the relationships between car traffic and alternative transport modes before and after the streetcar was reinstated.
+The `model_before` sets the baseline, showcasing the original dynamics between cars and other transport forms.
+The `model_after` reveals the traffic landscape streetcar service has resumed.
+This offers a comparative view to measure the efficacy of these public transit access improvements in mitigating car congestion.
 
-Together with a temporal trend, the `model_before` captures the baseline relationship between automobile traffic and other modes of transportation.
-As an illustration of possible changes in traffic patterns brought about by the improved streetcar service, the `model_after`, on the other hand, depicts the altered dynamics following the enhancement.
-By use of comparative analysis, these models demonstrate the effectiveness of the Route 508 enhancements in mitigating car congestion.
+This modeling approach not only provides insights into the immediate effects of transit policy changes but also establishes a methodological framework for future studies examining the longitudinal impact of urban transit interventions.
 
@@
 
-```{r residuals-before-model, fig.cap="Residuals of the negative binomial regression model before and after the transit improvement.
-This plot helps to visualize if there are any patterns or systematic deviations unexplained by the model.", echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
-# Read and prepare data
-analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
+```{r}
+#| label: fig-residualsmodel
+#| fig-cap: Residuals of the negative binomial regression model before and after the transit improvement.
+This plot helps to visualize if there are any patterns or systematic deviations unexplained by the model.
+#| echo: FALSE
+#| warning: FALSE
+#| message: FALSE
 
@@
 # Visualizing residuals to identify patterns
-par(mfrow = c(1, 2))
-plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
-plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")
+#par(mfrow = c(1, 2))
+#plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
+#plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")
 
-# Examining the normality of residuals
-# hist(resid(model_before), main = "Histogram of Residuals - Before")
-# hist(resid(model_after), main = "Histogram of Residuals - After")
+# Adjust overall plot margins (bottom, left, top, right) and spacing for multi-figure plots
+par(mfrow = c(1, 2), mar = c(5, 4, 4.5, 2) + 0.1, mgp = c(2, 0.7, 0), tcl = -0.3, cex.main = 0.9)
+
+# Visualizing residuals for the 'before improvement' model
+plot(residuals(model_before), 
+     type = "p", 
+     main = "Residuals Before Improvement", 
+     xlab = "Index", 
+     ylab = "Residuals", 
+     xlim = c(0, max(length(residuals(model_before)), length(residuals(model_after)))),
+     ylim = range(c(residuals(model_before), residuals(model_after))))
+
+# Visualizing residuals for the 'after improvement' model
+plot(residuals(model_after), 
+     type = "p", 
+     main = "Residuals After Improvement", 
+     xlab = "Index", 
+     ylab = "Residuals", 
+     xlim = c(0, max(length(residuals(model_before)), length(residuals(model_after)))),
+     ylim = range(c(residuals(model_before), residuals(model_after))))
+
 ```
 
-Figures 6 and 7 show an examination of residuals from the negative binomial regression models before and after the public transit improvement.
-For the model before improvements, the residuals plot does not indicate any apparent patterns or systemic deviations, suggesting that the model's assumptions may be adequate for the data.
-After the transit improvements, the residuals' plot again shows no obvious patterns, which suggests a random scatter.
+The residual plots depicted in @fig-residualsmodel serve as a diagnostic tool to evaluate the adequacy of the negative binomial regression models used in this study, both before and after the implementation of public transit improvements.
+The left panel of the figure illustrates the residuals for the model before the reinstatement of the Route 508 Lake Shore streetcar, showing a random dispersion of data points around the zero line.
+This lack of systematic pattern suggests that the model assumptions are met, indicating a good fit for the pre-improvement traffic data.
 
-\newpage
+Similarly, the right panel of the figure, representing the residuals post-improvement, exhibits a scatter of data points without discernible trends or biases.
+The randomness observed in the distribution of these residuals supports the conclusion that the post-improvement model is also well-fitted to the data, capturing the essential variability in daily car counts without leaving unexplained structure in the residuals.
 
-```{r model summaries, echo=FALSE, message=FALSE, warning=FALSE}
+```{r}
+#| echo: FALSE
+#| warning: FALSE
+#| message: FALSE
 
 # Read and prepare data
-analysis_data <- read.csv(here("data", "analysis_data", "cleaned-data.csv"))
 analysis_data$count_date <- as.Date(analysis_data$count_date)
@@
   mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))
+```
+
+```{r}
+#| label: tbl-modelbeforesummaries
+#| tbl-cap: "Model Summary of Traffic Patterns Before Improvement on the 508 Lakeshore Route"
+#| echo: FALSE
+#| warning: FALSE
+#| message: FALSE
 
@@
 model_before <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = before_improvement)
+
+# Tidying the model summaries 
+tidy_model_before <- broom::tidy(model_before)
+
+# Model before improvement table
+kable(tidy_model_before, format = "latex", booktabs = TRUE) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
+  column_spec(1, bold = TRUE, width = "3cm") %>%
+  column_spec(2, color = "blue", width = "3cm") %>%
+  column_spec(3, width = "3cm") %>%
+  column_spec(4, width = "3cm") %>%
+  column_spec(5, width = "3cm")
+
+```
+
+The pre-improvement model, as detailed in @tbl-modelbeforesummaries, reveals significant coefficients for both the intercept and daily bus traffic, alongside a temporal trend.
+With a coefficient of 9.34, the model showcases a high baseline volume of car traffic, independent of other variables.
+Bus traffic, represented by a coefficient of 0.261, appears to move in tandem with car traffic, implying that increases in bus usage were historically linked to increases in car traffic.
+However, the coefficients for pedestrian and bicycle traffic do not present a significant association with car traffic, reflecting a p-value greater than 0.05 for both.
+
+```{r}
+#| label: tbl-modelaftersummaries
+#| tbl-cap: "Model Summary of Traffic Patterns After Improvement on the 508 Lakeshore Route"
+#| echo: FALSE
+#| warning: FALSE
+#| message: FALSE
+
+# Building negative binomial regression models for before and after improvement
 model_after <- glm.nb(daily_cars ~ daily_bus + daily_peds + daily_bike + time_trend, data = after_improvement)
@@
 # Tidying the model summaries 
-tidy_model_before <- tidy(model_before)
-tidy_model_after <- tidy(model_after)
+tidy_model_after <- broom::tidy(model_after)
 
-# Model before improvement
-kable(tidy_model_before, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - Before Improvement") %>%
-  kable_styling(bootstrap_options = c("striped", "hover"),
-                full_width = F) %>%
+# Model after improvement table
+kable(tidy_model_after, format = "latex", booktabs = TRUE) %>%
+  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
   column_spec(1, bold = TRUE, width = "3cm") %>%
-  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
-  column_spec(3, width = "3cm")
-
-kable(tidy_model_after, format = "latex", booktabs = TRUE, caption = "Negative Binomial Regression Model - After Improvement") %>%
-  kable_styling(bootstrap_options = c("striped", "hover"),
-                full_width = F) %>%
-  column_spec(1, bold = TRUE, width = "3cm") %>%
-  column_spec(2, bold = TRUE, color = "blue", width = "3cm") %>%
-  column_spec(3, width = "3cm")
+  column_spec(2, color = "blue", width = "3cm") %>%
+  column_spec(3, width = "3cm") %>%
+  column_spec(4, width = "3cm") %>%
+  column_spec(5, width = "3cm")
 
@@
 
-Table 2 represents the traffic patterns before the improvement.
-These results produced significant coefficients for both the intercept and daily bus traffic, as well as the time trend variable.
-The intercept, representing the baseline log-count of cars when all other variables are zero, was significant with an estimate of 9.340 and a standard error of 0.072, indicating a high baseline traffic volume.
-Daily bus traffic had a positive and significant relationship with car traffic, with a coefficient of 0.261, suggesting that increases in bus traffic were associated with increases in car traffic.
-The time trend also showed a significant, albeit small, positive effect on car traffic `(β = 0.000287)`, indicating a gradual increase in car traffic over time.
+Conversely, the post-improvement model shown in @tbl-modelaftersummaries exhibits a marked change in these dynamics.
+The increase in the intercept to 12.283 suggests an elevated baseline of car traffic, potentially as a result of external influences not accounted for in the model.
+The coefficient for daily bus traffic experiences a considerable reduction to 0.096, signaling a weakened relationship between bus and car traffic in the wake of the transit improvements.
+This could indicate a disconnect of these modes of transport, possibly due to more passengers opting for the improved public transit system over personal vehicles.
+The time trend coefficient inverses to -0.001033, denoting a statistically significant decline in car traffic over time post-improvement, showcasing the potential efficacy of the public transit enhancements in reducing car traffic.
 
-Pedestrian and bicycle traffic, however, were not significantly associated with car traffic before the improvements (p \> 0.05 for both).
-
-Post-improvement, the model exhibits significant changes in the dynamics of traffic patterns.
-The intercept remains significant but increases to 12.283, reflecting an overall higher baseline of car traffic possibly due to other factors.
-The coefficient for daily bus traffic notably decreases to 0.096, indicating a diminished relationship between bus and car traffic after the public transit improvements.
-
-The time trend shifts from a slight increase to a significant decrease `(β = -0.001033)`, reflecting a reversal in the car traffic trend, which now shows a significant decline over time post-improvement.
-
-Variables for pedestrian and bicycle traffic remained non-significant, suggesting that these modes of transport did not have a substantial impact on car traffic volumes in either model.
-
-The model results indicate a significant reduction in the relationship between bus traffic and car congestion following the transit improvements, as shown by the decrease in the coefficient of daily bus traffic from 0.261 to 0.096.
-This suggests that improvements in the public transit system might be effectively alleviating car congestion.
-Additionally, the significant negative time trend in the post-improvement model underscores a successful overall reduction in car traffic between 2012 to 2020.
+These findings show that following the public transit improvements, the relationship between bus traffic and car congestion significantly diminished, potentially reflective of the transit system's capacity to alleviate car congestion.
+Additionally, the persistent non-significance of pedestrian and bicycle traffic variables implies these modes have not impacted car traffic volumes in the models considered.
 
@@
 
-These local findings align with research conducted by Verbavatz and Barthelemy `source` in which they found that urban area size and public transport density are pivotal in managing car traffic in cities, rather than urban density alone.
+These local findings align with research conducted by @Verbavatz2019 in which they found that urban area size and public transport density are pivotal in managing car traffic in cities, rather than urban density alone.
 The article further states that enhancing the density and accessibility of public transport is a viable path to mitigating traffic and its associated environmental impacts.
@@
 Moreover, the benefits of improved public transit does not stop at decreased car congestion.
-Insights from the American Public Transportation Association, which illustrate that public transit investments yield substantial congestion relief and economic benefits.
-Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested `source`.
+Insights from the @APTA2023, which illustrate that public transit investments yield substantial congestion relief and economic benefits.
+Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested .
 While on the individual level, households in America can save more than \$13,000 per year by using public transit and owning one less car.
 
-These advantages of public transportation investments are in alignment with the principles of sustainable development as defined by the UN's World Commission on Environment and Development: to meet current needs without compromising the ability of future generations to meet their own needs `source`.
+These advantages of public transportation investments are in alignment with the principles of sustainable development as defined by the UN's World Commission on Environment and Development: to meet current needs without compromising the ability of future generations to meet their own needs (@Lachman2013Chapter3).
 Together, these strands of evidence support the case for enhancing public transit as a strategic approach to improving urban traffic conditions and achieving broader sustainability goals.
@@
 
-Supporting this claim, Lexer Quamie's research (2011) `source` highlights the integral role of transportation policy in civil rights, particularly in periods of pronounced unemployment and income disparity.
+Supporting this claim, @Quamie2011 research highlights the integral role of transportation policy in civil rights, particularly in periods of pronounced unemployment and income disparity.
 Quamie emphasizes that transportation decisions significantly affect who can access employment opportunities and essential services, positioning public transit as a pivotal tool in the fight for civil rights.
@@
 
-# Additional data details
+## Motivation
 
-# Model details {#sec-model-details}
+For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.
 
-## Posterior predictive check
+-   The dataset was created to explore the impact of public transit infrastructure enhancements in Toronto, with a particular focus on the Route 508 Lake Shore streetcar.
+It hopes to address the gap in understanding how access to improved public transit could influence traffic congestion in urban environments.
 
-## Diagnostics
+Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?
+
+-   The dataset was compiled by Maria Mangru using data provided by the The City of Toronto's Transportation Services Division.
+
+Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.
+
+-   Unknown. The dataset was sourced from the City of Toronto's Transportation Services Division, which is typically funded by municipal budgets rather than specific grants.
+
+Any other comments?
+
+-   No additional comments.
+
+## Composition
+
+What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)?
+Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)?
+Please provide a description.
+
+-   Each instance within the dataset represents a specific time and location intersection within Toronto, capturing various modes of traffic volume, such as cars, buses, pedestrians, and cyclists.
+
+How many instances are there in total (of each type, if appropriate)?
+
+-   Over 200 instances in the cleaned dataset.
+
+Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set?
+If the dataset is a sample, then what is the larger set?
+Is the sample representative of the larger set (for example, geographic coverage)?
+If so, please describe how this representativeness was validated/verified.
+If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).
+
+-   The dataset is intended to represent all traffic volume counts at selected city intersections and is not a random sample; it's a comprehensive aggregation across multiple years.
+
+What data does each instance consist of?
+“Raw” data (for example, unprocessed text or images) or features?
+In either case, please provide a description.
+Each instance consists of counts of death including demographic data (race and sex) and year.
+Is there a label or target associated with each instance?
+If so, please provide a description.
+
+-   Each instance includes aggregated traffic volume counts for different modes of transport at various intersections in Toronto in addition to the date this instance was observed.
+
+Is there a label or target associated with each instance?
+
+-   Yes, instances could be labeled with unique identifiers such as the intersection ID and the date and time of the traffic counts.
+
+Is any information missing from individual instances?
+If so, please provide a description, explaining why this information is missing (for example, because it was unavailable).
+This does not include intentionally removed information, but might include, for example, redacted text.
+
+-   No.
+
+Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)? If so, please describe how these relationships are made explicit.
+
+-   The data includes both temporal and spatial data for each instance such as the time and location coordinates.
+
+Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.
+
+-   No.
+
+Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.
+
+-   As with any real-world dataset, some level of noise, error, or redundancy is likely but none have been observed to date.
+
+Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)?
+If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer?
+Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.
+
+-   The dataset is self-contained and does not rely on external resources for its primary analyses.
+
+Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals’ non-public communications)?
+If so, please provide a description.
+
+-   No, the dataset consists of aggregated traffic counts that are non-confidential and derived from public domain sources.
+
+Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.
+
+-   No, the dataset is strictly numerical and traffic-related, with no content that could be considered offensive or anxiety-inducing.
+
+Does the dataset identify any sub-populations (for example, by age, gender)?
+If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.
+
+-   No, the dataset does not identify any sub-populations, as it deals with vehicles and traffic, not individuals.
+
+Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.
+
+-   No, individuals cannot be identified from this dataset as it does not contain personal data.
+
+Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)?
+If so, please provide a description.
+
+-   No sensitive information pertaining to individuals is included. The dataset only contains information related to traffic volumes and patterns.
+
+Any other comments?
+
+-   No additional comments.
+
+## Collection process
+
+How was the data associated with each instance acquired?
+Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)?
+If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified?
+If so, please describe how.
+
+-   The data for each instance were obtained from automatic traffic recorders and manual counts at intersections. These data collection methods are considered direct observations of traffic volumes.
+
+What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)?
+How were these mechanisms or procedures validated?
+
+-   Data were collected using a combination of hardware apparatuses, such as automatic traffic recorders (ATR), and manual turning movement counts (TMCs).
+The procedures and equipment were likely standardized and validated by the city's transportation services to ensure accuracy.
+
+If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?
+
+-   The dataset is a comprehensive collection of traffic volumes at various intersections within Toronto, not a sample from a larger set.
+
+Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?
+
+-   The data collection was performed by the City of Toronto's Transportation Services Division, which employs trained personnel for such tasks.
+Compensation details for city employees are typically determined by municipal guidelines and salary scales.
+
+Over what timeframe was the data collected?
+
+-   The dataset encompasses traffic volume data from 2010 to 2024, which likely matches the creation timeframe for the instances.
+
+Were any ethical review processes conducted (for example, by an institutional review board)?
+If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.
+
+-   For data collection that involves public transportation metrics without personal identifiers, ethical review processes are generally not applicable.
+
+Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?
+
+-   Data were collected directly from the traffic volumes and movements at intersections, not from individuals.
+
+Were the individuals in question notified about the data collection?
+If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.
+
+-   Not applicable as the dataset consists of traffic counts, not personal data.
+
+Did the individuals in question consent to the collection and use of their data?
+
+-   Not applicable as no individual personal data was collected.
+
+If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses?
+
+-   Consent and mechanisms to revoke consent are not applicable to this dataset.
+
+Has an analysis of the potential impact of the dataset and its use on data subjects been conducted
+
+-   Given the dataset contains aggregated non-personal data regarding traffic volumes, an analysis of the impact on data subjects would not be applicable.
+
+Any other comments?
+
+-   No additional comments. The dataset is designed for urban planning and transportation analysis and does not involve personal data collection.
+
+## Preprocessing/cleaning/labeling
+
+Was any preprocessing/cleaning/labeling of the data done?
+
+-   Yes, preprocessing and cleaning of the data were done.
+This included aggregating traffic volumes to daily totals, cleaning anomalies, filtering data for relevant time periods, and aligning the dataset for comparative analysis before and after the implementation of transit improvements.
+
+Was the “raw” data saved in addition to the preprocessed/cleaned/labeled data?
+
+-   Yes. Both the raw and cleaned data were saved to this project.
+
+Is the software that was used to preprocess/clean/label the data available?
+
+-   Yes, R programming language and its various packages were used for data preprocessing, cleaning, and analysis.
+The specific scripts used for this can be found on this GitHub repository (<https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline>).
+
+Any other comments?
+
+-   No additional comments.
+
+## Uses
+
+Has the dataset been used for any tasks already?
+
+-   The dataset has been used for the current study, which investigates the impact of public transit improvements on urban car congestion, utilizing traffic volume data to examine changes in traffic patterns over time.
+
+Is there a repository that links to any or all papers or systems that use the dataset?
+
+-   This repository (<https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline>) is the only one to day which uses this dataset.
+
+What (other) tasks could the dataset be used for?
+
+-   Beyond studying traffic congestion, the dataset could potentially be used for urban planning, environmental impact assessments, and public policy formulation regarding transportation and mobility in urban areas.
+It may also be relevant for studies on the socio-economic implications of traffic patterns.
+
+Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?
+
+-   The dataset is specific to traffic volumes in Toronto and reflects the unique urban and transit infrastructure of the city.
+Users should be cautious about generalizing findings to other cities without considering local differences.
+
+Are there tasks for which the dataset should not be used?
+
+-   The dataset should not be used for tasks that require personal, sensitive, or confidential data as it is composed of aggregated non-personal traffic volumes.
+
+Any other comments?
+
+-   No additional comments.
+
+## Distribution
+
+Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created?
+If so, please provide a description.
+
+-   The dataset is available through the NYC Open Data Profile.
+
+When will the dataset be distributed?
+
+-   The dataset is uploaded and distributed annually through the NYC Open Data Profile.
+
+Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
+If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.
+
+-   None that are known.
+
+Do any export controls or other regulatory restrictions apply to the dataset or to individual instances If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.
+
+-   None that are known.
+
+Will the dataset be distributed to third parties outside of the entity on behalf of which the dataset was created?
+
+-   The dataset, as obtained from the Toronto OpenData portal, is already publicly available. As such, it can be accessed by third parties for analysis and research purposes.
+
+How will the dataset be distributed? Does the dataset have a digital object identifier (DOI)?
+
+-   The dataset is distributed through the Toronto OpenData portal, typically in CSV format.
+It does not have a DOI as it is continuously updated by the city's transportation services division and not a static dataset.
+
+When will the dataset be distributed?
+
+-   The dataset is currently available and is periodically updated by the Toronto OpenData Portal.
+
+Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)?
+
+-   The dataset from the Toronto OpenData portal is under an open government license, which allows for free use, modification, and sharing of the data, as long as the City of Toronto is credited for the data.
+
+Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
+
+-   There are no known IP-based or other restrictions imposed by third parties on the dataset.
+
+Do any export controls or other regulatory restrictions apply to the dataset or to individual instances?
+
+-   There are no known export controls or regulatory restrictions that apply to this dataset, as it consists of non-sensitive, aggregated traffic volume data.
+
+Any other comments?
+
+-   Any researcher or user of this dataset should ensure proper attribution as per the licensing terms of the City of Toronto and should verify that they are using the most up-to-date data available.
+
+## Maintenance
+
+Who will be supporting/hosting/maintaining the dataset?
+
+-   The dataset is maintained by the City of Toronto's Transportation Services Division.
+
+How can the owner/curator/manager of the dataset be contacted?
+
+-   Inquiries regarding the dataset can be directed to the City of Toronto's Transportation Services team at (VolumeCollision_Requests\@toronto.ca).
+
+Is there an erratum? There is no official erratum; however, the dataset is subject to updates and corrections as part of its regular maintenance schedule.
+
+Will the dataset be updated?
+
+-   The dataset is regularly updated by the Transportation Services Division. Updates, including changes in traffic volumes or infrastructure, are typically communicated through the OpenData portal.
+
+Are there applicable limits on the retention of the data associated with the instances?
+
+-   The dataset does not include personal data, and therefore retention limits associated with personal data do not apply.
+
+Will older versions of the dataset continue to be supported/hosted/maintained?
+
+-   The OpenData portal maintains a history of updates, allowing for access to previous versions of the dataset. However, users are encouraged to use the most current data.
+
+If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?
+
+-   Contributions to the city's open data, such as corrections or enhancements, can be done through communication with the Open Data team.
+However, contributions are subject to validation and incorporation by the city staff.
+
+Any other comments?
+
+-   Users of the dataset should regularly check the OpenData portal for updates and note the date of the version they are using in their analysis to ensure reproducibility.
 

### 2024-04-18 scripts/03-exploring_cleaned_data.R

@@ -17,12 +17,12 @@ library(here)
 ### Creating Summary Statistics and Visualizations ###
 
 # Load the cleaned data
-cleaned_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
+analysis_data <- read_csv(here("data", "analysis_data", "cleaned-data.csv"))
 
 # Summary Statistics
 
 # Generate and view summary statistics for traffic data
-summary_stats <- cleaned_data %>%
+summary_stats <- analysis_data %>%
   summarise(
     Mean_Cars = mean(daily_cars, na.rm = TRUE),
     Median_Cars = median(daily_cars, na.rm = TRUE),
@@ -39,7 +39,7 @@ View(summary_stats)
 # Visualize Data
 
 # Visualizing daily car traffic over time
-line_graph_car <- cleaned_data %>%
+line_graph_car <- analysis_data %>%
   ggplot(aes(x = count_date, y = daily_cars)) +
   geom_smooth(se = FALSE) +  
   labs(title = "Daily Cars Traffic Over Time",
@@ -54,7 +54,7 @@ line_graph_car
 ggsave("./other/outputs/car_traffic_over_time.png", plot = line_graph_car, width = 12, height = 6, units = "in")
 
 # Visualizing daily bus traffic over time
-line_graph_buses <- cleaned_data %>%
+line_graph_buses <- analysis_data %>%
   ggplot(aes(x = count_date, y = daily_bus)) +
   geom_smooth(se = FALSE) +
   labs(title = "Daily Bus Traffic Over Time",
@@ -69,7 +69,7 @@ line_graph_buses
 ggsave("./other/outputs/bus_traffic_over_time.png", plot = line_graph_buses, width = 12, height = 6, units = "in")
 
 # Visualizing daily pedestrian traffic over time
-line_graph_pedestrians <- cleaned_data %>%
+line_graph_pedestrians <- analysis_data %>%
   ggplot(aes(x = count_date, y = daily_peds)) +
   geom_smooth(se = FALSE) +
   labs(title = "Daily Pedestrian Traffic Over Time",
@@ -84,7 +84,7 @@ line_graph_pedestrians
 ggsave("./other/outputs/pedestrian_traffic_over_time.png", plot = line_graph_pedestrians, width = 12, height = 6, units = "in")
 
 # Visualizing daily bike traffic over time
-line_graph_cyclists <- cleaned_data %>%
[…]

### 2024-04-19 paper/datasheet_template.qmd
updated datasheet

---
title: "Datasheet for 'A dataset'"
subtitle: "My subtitle if needed"
author: 
  - First author
  - Another author
thanks: "Code and data are available at: LINK."
date: today
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
  -   The dataset was created to explore the impact of public transit infrastructure enhancements in Toronto, with a particular focus on the Route 508 Lake Shore streetcar. It hopes to address the gap in understanding how access to improved public transit could influence traffic congestion in urban environments.
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
  -   The dataset was compiled by Maria Mangru using data provided by the The City of Toronto's Transportation Services Division.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
  -   Unknown. The dataset was sourced from the City of Toronto's Transportation Services Division, which is typically funded by municipal budgets rather than specific grants.
4. *Any other comments?*
  -   No additional comments.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.* 
  -   Each instance within the dataset represents a specific time and location intersection within Toronto, capturing various modes of traffic volume, such as cars, buses, pedestrians, and cyclists.

2. *How many instances are there in total (of each type, if appropriate)?*
  -   Over 200 instances in the cleaned dataset.

3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
  -   The dataset is intended to represent all traffic volume counts at selected city intersections and is not a random sample; it's a comprehensive aggregation across multiple years.

4. *What data does each instance consist of? “Raw” data (for example, unprocessed text or images) or features? In either case, please provide a description. Each instance consists of counts of death including demographic data (race and sex) and year. Is there a label or target associated with each instance? If so, please provide a description.* 
  -   Each instance includes aggregated traffic volume counts for different modes of transport at various intersections in Toronto in addition to the date this instance was observed.

5. *Is there a label or target associated with each instance?*
  -   Yes, instances could be labeled with unique identifiers such as the intersection ID and the date and time of the traffic counts.

6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.* 
  -   No.

7. *Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)? If so, please describe how these relationships are made explicit.* 
  -   The data includes both temporal and spatial data for each instance such as the time and location coordinates.

8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
  -   No.

9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
  -   As with any real-world dataset, some level of noise, error, or redundancy is likely but none have been observed to date.

10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
  -   The dataset is self-contained and does not rely on external resources for its primary analyses.

11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals’ non-public communications)? If so, please provide a description.*
  -   No, the dataset consists of aggregated traffic counts that are non-confidential and derived from public domain sources.

12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
  -   No, the dataset is strictly numerical and traffic-related, with no content that could be considered offensive or anxiety-inducing.

13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
  -   No, the dataset does not identify any sub-populations, as it deals with vehicles and traffic, not individuals.

14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
  -   No, individuals cannot be identified from this dataset as it does not contain personal data.

15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
  -   No sensitive information pertaining to individuals is included. The dataset only contains information related to traffic volumes and patterns.

16. *Any other comments?*
-   No additional comments.

**Collection process**
1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
  -   The data for each instance were obtained from automatic traffic recorders and manual counts at intersections. These data collection methods are considered direct observations of traffic volumes.

2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
  -   Data were collected using a combination of hardware apparatuses, such as automatic traffic recorders (ATR), and manual turning movement counts (TMCs). The procedures and equipment were likely standardized and validated by the city's transportation services to ensure accuracy.

3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
  -   The dataset is a comprehensive collection of traffic volumes at various intersections within Toronto, not a sample from a larger set.

4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
  -   The data collection was performed by the City of Toronto's Transportation Services Division, which employs trained personnel for such tasks. Compensation details for city employees are typically determined by municipal guidelines and salary scales.

5. *Over what timeframe was the data collected?*
  -   The dataset encompasses traffic volume data from 2010 to 2024, which likely matches the creation timeframe for the instances.

6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
  -   For data collection that involves public transportation metrics without personal identifiers, ethical review processes are generally not applicable.

7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
  -   Data were collected directly from the traffic volumes and movements at intersections, not from individuals.

8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
  -   Not applicable as the dataset consists of traffic counts, not personal data.

9. *Did the individuals in question consent to the collection and use of their data?*
  -   Not applicable as no individual personal data was collected.

10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses?*
  -   Consent and mechanisms to revoke consent are not applicable to this dataset.

11. *Has an analysis of the potential impact of the dataset and its use on data subjects been conducted*
  -   Given the dataset contains aggregated non-personal data regarding traffic volumes, an analysis of the impact on data subjects would not be applicable.

12. *Any other comments?*
  -   No additional comments. The dataset is designed for urban planning and transportation analysis and does not involve personal data collection.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done?*
  -   Yes, preprocessing and cleaning of the data were done. This included aggregating traffic volumes to daily totals, cleaning anomalies, filtering data for relevant time periods, and aligning the dataset for comparative analysis before and after the implementation of transit improvements.

2. *Was the “raw” data saved in addition to the preprocessed/cleaned/labeled data?*
  -   Yes. Both the raw and cleaned data were saved to this project.

3. *Is the software that was used to preprocess/clean/label the data available?*
  -   Yes, R programming language and its various packages were used for data preprocessing, cleaning, and analysis. The specific scripts used for this can be found on this GitHub repository (<https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline>).

4. *Any other comments?*
  -   No additional comments.

**Uses**

1. *Has the dataset been used for any tasks already?*
  -   The dataset has been used for the current study, which investigates the impact of public transit improvements on urban car congestion, utilizing traffic volume data to examine changes in traffic patterns over time.

2. *Is there a repository that links to any or all papers or systems that use the dataset?*
  -   This repository (<https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline>) is the only one to day which uses this dataset.

3. *What (other) tasks could the dataset be used for?*
  -   Beyond studying traffic congestion, the dataset could potentially be used for urban planning, environmental impact assessments, and public policy formulation regarding transportation and mobility in urban areas. It may also be relevant for studies on the socio-economic implications of traffic patterns.

4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?*
  -   The dataset is specific to traffic volumes in Toronto and reflects the unique urban and transit infrastructure of the city. Users should be cautious about generalizing findings to other cities without considering local differences.

5. *Are there tasks for which the dataset should not be used?*
  -   The dataset should not be used for tasks that require personal, sensitive, or confidential data as it is composed of aggregated non-personal traffic volumes.

6. *Any other comments?*
  -   No additional comments.

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
  -   The dataset is available through the NYC Open Data Profile.

2. *When will the dataset be distributed?*
  -   The dataset is uploaded and distributed annually through the NYC Open Data Profile.

3. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
  -   None that are known.

4. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
  -   None that are known.

5. *Will the dataset be distributed to third parties outside of the entity on behalf of which the dataset was created?*
  -   The dataset, as obtained from the Toronto OpenData portal, is already publicly available. As such, it can be accessed by third parties for analysis and research purposes.

6. *How will the dataset be distributed? Does the dataset have a digital object identifier (DOI)?*
  -   The dataset is distributed through the Toronto OpenData portal, typically in CSV format. It does not have a DOI as it is continuously updated by the city's transportation services division and not a static dataset.

7. *When will the dataset be distributed?*
  -   The dataset is currently available and is periodically updated by the Toronto OpenData Portal.

8. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)?*
  -   The dataset from the Toronto OpenData portal is under an open government license, which allows for free use, modification, and sharing of the data, as long as the City of Toronto is credited for the data.

9. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances?*
  -   There are no known IP-based or other restrictions imposed by third parties on the dataset.

10. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances?*
  -   There are no known export controls or regulatory restrictions that apply to this dataset, as it consists of non-sensitive, aggregated traffic volume data.

11. *Any other comments?*
  -   Any researcher or user of this dataset should ensure proper attribution as per the licensing terms of the City of Toronto and should verify that they are using the most up-to-date data available.

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
  -   The dataset is maintained by the City of Toronto's Transportation Services Division.

2. *How can the owner/curator/manager of the dataset be contacted?*
  -   Inquiries regarding the dataset can be directed to the City of Toronto's Transportation Services team at (VolumeCollision_Requests\@toronto.ca).

3. *Is there an erratum?*
  - There is no official erratum; however, the dataset is subject to updates and corrections as part of its regular maintenance schedule.

4. *Will the dataset be updated?*
  -   The dataset is regularly updated by the Transportation Services Division. Updates, including changes in traffic volumes or infrastructure, are typically communicated through the OpenData portal.

5. *Are there applicable limits on the retention of the data associated with the instances?*
  -   The dataset does not include personal data, and therefore retention limits associated with personal data do not apply.

6. *Will older versions of the dataset continue to be supported/hosted/maintained?*
  -   The OpenData portal maintains a history of updates, allowing for access to previous versions of the dataset. However, users are encouraged to use the most current data.

7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?*
  -   Contributions to the city's open data, such as corrections or enhancements, can be done through communication with the Open Data team. However, contributions are subject to validation and incorporation by the city staff.

8. *Any other comments?*
  -   Users of the dataset should regularly check the OpenData portal for updates and note the date of the version they are using in their analysis to ensure reproducibility.

\newpage

# References

### 2024-04-19 paper/paper.qmd
updated datasheet

@@
 
-\appendix
-
-# Appendix {.unnumbered}
-
-## Motivation
-
-For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.
-
--   The dataset was created to explore the impact of public transit infrastructure enhancements in Toronto, with a particular focus on the Route 508 Lake Shore streetcar.
-It hopes to address the gap in understanding how access to improved public transit could influence traffic congestion in urban environments.
-
-Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?
-
--   The dataset was compiled by Maria Mangru using data provided by the The City of Toronto's Transportation Services Division.
-
-Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.
-
--   Unknown. The dataset was sourced from the City of Toronto's Transportation Services Division, which is typically funded by municipal budgets rather than specific grants.
-
-Any other comments?
-
--   No additional comments.
-
-## Composition
-
-What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)?
-Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)?
-Please provide a description.
-
--   Each instance within the dataset represents a specific time and location intersection within Toronto, capturing various modes of traffic volume, such as cars, buses, pedestrians, and cyclists.
-
-How many instances are there in total (of each type, if appropriate)?
-
--   Over 200 instances in the cleaned dataset.
-
-Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set?
-If the dataset is a sample, then what is the larger set?
-Is the sample representative of the larger set (for example, geographic coverage)?
-If so, please describe how this representativeness was validated/verified.
-If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).
-
--   The dataset is intended to represent all traffic volume counts at selected city intersections and is not a random sample; it's a comprehensive aggregation across multiple years.
-
-What data does each instance consist of?
-“Raw” data (for example, unprocessed text or images) or features?
-In either case, please provide a description.
-Each instance consists of counts of death including demographic data (race and sex) and year.
-Is there a label or target associated with each instance?
-If so, please provide a description.
-
--   Each instance includes aggregated traffic volume counts for different modes of transport at various intersections in Toronto in addition to the date this instance was observed.
-
-Is there a label or target associated with each instance?
-
--   Yes, instances could be labeled with unique identifiers such as the intersection ID and the date and time of the traffic counts.
-
-Is any information missing from individual instances?
-If so, please provide a description, explaining why this information is missing (for example, because it was unavailable).
-This does not include intentionally removed information, but might include, for example, redacted text.
-
--   No.
-
-Are relationships between individual instances made explicit (for example, users’ movie ratings, social network links)? If so, please describe how these relationships are made explicit.
-
--   The data includes both temporal and spatial data for each instance such as the time and location coordinates.
-
-Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.
-
--   No.
-
-Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.
-
--   As with any real-world dataset, some level of noise, error, or redundancy is likely but none have been observed to date.
-
-Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)?
-If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer?
-Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.
-
--   The dataset is self-contained and does not rely on external resources for its primary analyses.
-
-Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals’ non-public communications)?
-If so, please provide a description.
-
--   No, the dataset consists of aggregated traffic counts that are non-confidential and derived from public domain sources.
-
-Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.
-
--   No, the dataset is strictly numerical and traffic-related, with no content that could be considered offensive or anxiety-inducing.
-
-Does the dataset identify any sub-populations (for example, by age, gender)?
-If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.
-
--   No, the dataset does not identify any sub-populations, as it deals with vehicles and traffic, not individuals.
-
-Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.
-
--   No, individuals cannot be identified from this dataset as it does not contain personal data.
-
-Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)?
-If so, please provide a description.
-
--   No sensitive information pertaining to individuals is included. The dataset only contains information related to traffic volumes and patterns.
-
-Any other comments?
-
--   No additional comments.
-
-## Collection process
-
-How was the data associated with each instance acquired?
-Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)?
-If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified?
-If so, please describe how.
-
--   The data for each instance were obtained from automatic traffic recorders and manual counts at intersections. These data collection methods are considered direct observations of traffic volumes.
-
-What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)?
-How were these mechanisms or procedures validated?
-
--   Data were collected using a combination of hardware apparatuses, such as automatic traffic recorders (ATR), and manual turning movement counts (TMCs).
-The procedures and equipment were likely standardized and validated by the city's transportation services to ensure accuracy.
-
-If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?
-
--   The dataset is a comprehensive collection of traffic volumes at various intersections within Toronto, not a sample from a larger set.
-
-Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?
-
--   The data collection was performed by the City of Toronto's Transportation Services Division, which employs trained personnel for such tasks.
-Compensation details for city employees are typically determined by municipal guidelines and salary scales.
-
-Over what timeframe was the data collected?
-
--   The dataset encompasses traffic volume data from 2010 to 2024, which likely matches the creation timeframe for the instances.
-
-Were any ethical review processes conducted (for example, by an institutional review board)?
-If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.
-
--   For data collection that involves public transportation metrics without personal identifiers, ethical review processes are generally not applicable.
-
-Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?
-
--   Data were collected directly from the traffic volumes and movements at intersections, not from individuals.
-
-Were the individuals in question notified about the data collection?
-If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.
-
--   Not applicable as the dataset consists of traffic counts, not personal data.
-
-Did the individuals in question consent to the collection and use of their data?
-
--   Not applicable as no individual personal data was collected.
-
-If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses?
-
--   Consent and mechanisms to revoke consent are not applicable to this dataset.
-
-Has an analysis of the potential impact of the dataset and its use on data subjects been conducted
-
--   Given the dataset contains aggregated non-personal data regarding traffic volumes, an analysis of the impact on data subjects would not be applicable.
-
-Any other comments?
-
--   No additional comments. The dataset is designed for urban planning and transportation analysis and does not involve personal data collection.
-
-## Preprocessing/cleaning/labeling
-
-Was any preprocessing/cleaning/labeling of the data done?
-
--   Yes, preprocessing and cleaning of the data were done.
-This included aggregating traffic volumes to daily totals, cleaning anomalies, filtering data for relevant time periods, and aligning the dataset for comparative analysis before and after the implementation of transit improvements.
-
-Was the “raw” data saved in addition to the preprocessed/cleaned/labeled data?
-
--   Yes. Both the raw and cleaned data were saved to this project.
-
-Is the software that was used to preprocess/clean/label the data available?
-
--   Yes, R programming language and its various packages were used for data preprocessing, cleaning, and analysis.
-The specific scripts used for this can be found on this GitHub repository (<https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline>).
-
-Any other comments?
-
--   No additional comments.
-
-## Uses
-
-Has the dataset been used for any tasks already?
-
--   The dataset has been used for the current study, which investigates the impact of public transit improvements on urban car congestion, utilizing traffic volume data to examine changes in traffic patterns over time.
-
-Is there a repository that links to any or all papers or systems that use the dataset?
-
--   This repository (<https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline>) is the only one to day which uses this dataset.
-
-What (other) tasks could the dataset be used for?
-
--   Beyond studying traffic congestion, the dataset could potentially be used for urban planning, environmental impact assessments, and public policy formulation regarding transportation and mobility in urban areas.
-It may also be relevant for studies on the socio-economic implications of traffic patterns.
-
-Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses?
-
--   The dataset is specific to traffic volumes in Toronto and reflects the unique urban and transit infrastructure of the city.
-Users should be cautious about generalizing findings to other cities without considering local differences.
-
-Are there tasks for which the dataset should not be used?
-
--   The dataset should not be used for tasks that require personal, sensitive, or confidential data as it is composed of aggregated non-personal traffic volumes.
-
-Any other comments?
-
--   No additional comments.
-
-## Distribution
-
-Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created?
-If so, please provide a description.
-
--   The dataset is available through the NYC Open Data Profile.
-
-When will the dataset be distributed?
-
--   The dataset is uploaded and distributed annually through the NYC Open Data Profile.
-
-Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
-If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.
-
--   None that are known.
-
-Do any export controls or other regulatory restrictions apply to the dataset or to individual instances If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.
-
--   None that are known.
-
-Will the dataset be distributed to third parties outside of the entity on behalf of which the dataset was created?
-
--   The dataset, as obtained from the Toronto OpenData portal, is already publicly available. As such, it can be accessed by third parties for analysis and research purposes.
-
-How will the dataset be distributed? Does the dataset have a digital object identifier (DOI)?
-
--   The dataset is distributed through the Toronto OpenData portal, typically in CSV format.
-It does not have a DOI as it is continuously updated by the city's transportation services division and not a static dataset.
-
-When will the dataset be distributed?
-
--   The dataset is currently available and is periodically updated by the Toronto OpenData Portal.
-
-Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)?
-
--   The dataset from the Toronto OpenData portal is under an open government license, which allows for free use, modification, and sharing of the data, as long as the City of Toronto is credited for the data.
-
-Have any third parties imposed IP-based or other restrictions on the data associated with the instances?
-
--   There are no known IP-based or other restrictions imposed by third parties on the dataset.
-
-Do any export controls or other regulatory restrictions apply to the dataset or to individual instances?
-
--   There are no known export controls or regulatory restrictions that apply to this dataset, as it consists of non-sensitive, aggregated traffic volume data.
-
-Any other comments?
-
--   Any researcher or user of this dataset should ensure proper attribution as per the licensing terms of the City of Toronto and should verify that they are using the most up-to-date data available.
-
-## Maintenance
-
-Who will be supporting/hosting/maintaining the dataset?
-
--   The dataset is maintained by the City of Toronto's Transportation Services Division.
-
-How can the owner/curator/manager of the dataset be contacted?
-
--   Inquiries regarding the dataset can be directed to the City of Toronto's Transportation Services team at (VolumeCollision_Requests\@toronto.ca).
-
-Is there an erratum? There is no official erratum; however, the dataset is subject to updates and corrections as part of its regular maintenance schedule.
-
-Will the dataset be updated?
-
--   The dataset is regularly updated by the Transportation Services Division. Updates, including changes in traffic volumes or infrastructure, are typically communicated through the OpenData portal.
-
-Are there applicable limits on the retention of the data associated with the instances?
-
--   The dataset does not include personal data, and therefore retention limits associated with personal data do not apply.
-
-Will older versions of the dataset continue to be supported/hosted/maintained?
-
--   The OpenData portal maintains a history of updates, allowing for access to previous versions of the dataset. However, users are encouraged to use the most current data.
-
-If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?
-
--   Contributions to the city's open data, such as corrections or enhancements, can be done through communication with the Open Data team.
-However, contributions are subject to validation and incorporation by the city staff.
-
-Any other comments?
-
--   Users of the dataset should regularly check the OpenData portal for updates and note the date of the version they are using in their analysis to ensure reproducibility.
-
-\newpage
-
 # References

### 2024-04-19 scripts/00-simulate_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Simulates the data set `Traffic Volumes at Intersections for All Modes`
 # Author: Maria Mangru
-# Date: March 2024
+# Date: 18th April 2024
 # Contact:maria.mangru@mail.utoronto.ca
 # License: MIT
 

### 2024-04-19 scripts/02-data_cleaning.R

@@ -12,6 +12,7 @@ library(readr)
 library(stringr)
 library(lubridate)
 library(here)
+library(arrow)  
 
 #### Clean data ####
 
@@ -101,5 +102,5 @@ daily_traffic <- daily_traffic %>%
 #### Save data ####
 write.csv(daily_traffic, here("data", "analysis_data", "cleaned-data.csv"), row.names = FALSE)
 
-  
-colnames(raw_data_2010_2019)
+# Saving as Parquet
+write_parquet(daily_traffic, here("data", "analysis_data", "cleaned-data.parquet"))

### 2024-04-19 scripts/03-exploring_cleaned_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Exploration of the dataset to gather trends and determine appropriate models  
 # Author: Maria Mangru
-# Date: March 2024
+# Date: 18th April 2024 
 # Contact:maria.mangru@mail.utoronto.ca
 # License: MIT
 

### 2024-04-19 scripts/04-test_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Tests simulate data, data cleaning and exploring cleaned data scripts
 # Author: Maria Mangru
-# Date: March 2024
+# Date: 18th April 2024
 # Contact:maria.mangru@mail.utoronto.ca
 # License: MIT
 

### 2024-04-19 paper/datasheet.qmd
final grammar edits to paper and placement of graphs

--- 2024-04-19 paper/datasheet_template.qmd
@@
 ---
-title: "Datasheet for 'A dataset'"
-subtitle: "My subtitle if needed"
+title: "Datasheet for 'Impact of Public Transit Infrastructure Access on Toronto's Urban Traffic Congestion'"
+subtitle: "Analyzing the effects of the 508 Lake Shore Streetcar improvements"
 author: 
-  - First author
-  - Another author
-thanks: "Code and data are available at: LINK."
+  - Maria Mangru
+thanks: "Code and data are available at: https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline."
 date: today
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This datasheet accompanies a dataset derived from the City of Toronto's Transportation Services Division, aimed at analyzing the influence of public transit access on urban traffic patterns.
+It encompasses data from 2010 to 2020, focusing on the periods before and after the reinstating of the 508 Lake Shore streetcar route.
+The analysis seeks to understand how improved public transit options can reduce urban congestion and contribute to city planning initiatives."
 format: pdf

### 2024-04-19 paper/paper.qmd
final grammar edits to paper and placement of graphs

@@
 subtitle: "Examining Traffic Patterns Before and After Public Transit Access was Reinstated"
-author: 
-  - Maria Mangru
-thanks: "Code and data are available at: https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline."
+author: Maria Mangru
+thanks: "Code and data are available at: [https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline](https://github.com/MariaMangru/Traffic-Congestion-on-Toronto-s-Shoreline)"
 date: today
@@
 abstract: "This study investigates the impact of access to the Route 508 Lake Shore streetcar in Toronto on reducing urban car congestion from 2011 to 2020.
-By analyzing traffic volume data from the City of Toronto's Transportation Services Division and employing negative binomial regression models, we observed a notable shift in travel behavior, highlighted by a significant decrease in the correlation between bus traffic and car congestion.
+By analyzing traffic volume data from the City of Toronto's Transportation Services Division and employing negative binomial regression models, we observed a notable shift in travel behaviour, highlighted by a significant decrease in the correlation between bus traffic and car congestion.
 Specifically, the coefficient for daily bus traffic reduced from 0.261 to 0.096 after access was reinstated, indicating a reduction in car traffic as more public transit options became available.
@@
 By employing traffic volume data collected by the City of Toronto's Transportation Services Division, we utilize negative binomial regression models to quantify the shifts in traffic patterns from 2011 to 2020.
-The estimand of this study is the change in the correlation between daily bus traffic and daily car traffic following the transit improvements.
-For the purposes of this research, the terms "improvements" and "reinstating of the streetcar" are used interchangeably.
+The estimand of this study is the change in the correlation between daily bus traffic and daily car traffic following transit improvements.
+For this research, the terms "improvements" and "reinstating of the streetcar" are used interchangeably.
 Both terms refer to the suite of upgrades and changes implemented along the 508 Lake Shore route, which included the restoration of streetcar services that had previously been discontinued or altered.'
@@
 
-This investigation not only sheds light on the specific case of the Route 508 Lake Shore but aims to contribute to a broader understanding of how enhancing public transit infrastructure can significantly alter urban mobility patterns.
+This investigation not only sheds light on the specific case of Route 508 Lake Shore but aims to contribute to a broader understanding of how enhancing public transit infrastructure can significantly alter urban mobility patterns.
 
@@
 Further, a suite of R packages, including several from the tidyverse collection were utilized.
-The @dplyr package was used the transformation and summarization of traffic data.
+The @dplyr package was used for the transformation and summarization of traffic data.
 For importing and processing string data, @readr and @stringr were utilized.
@@
 While the dataset was merged from two data files which span from 2010 to 2019 and 2020 to 2029, this research will focus on December 2011 to December 2014 and November 2018 to February 2020.
-These years were chosen to provide a comprehensive view of traffic patterns prior to, during, and following the enhancement of the 508 Lake Shore route.
+These years were chosen to provide a comprehensive view of traffic patterns before, during, and following the enhancement of the 508 Lake Shore route.
 The data collection period was intentionally concluded in February 2020 to ensure that the results were not influenced by the disruptions caused by the COVID-19 pandemic.
+
+\newpage
 
@@
 
-The summary statistics presented in @tbl-trafficstats offers a quantitative comparison across different modes of transportation, capturing the daily traffic volumes on the 508 Lake Shore route.
+The summary statistics presented in @tbl-trafficstats offer a quantitative comparison across different modes of transportation, capturing the daily traffic volumes on the 508 Lake Shore route.
 Notably, **Daily Car Traffic** significantly surpasses other modes of transit, with mean and median values at 14,169.85 and 13,291.5, respectively.
@@
 
-\newpage
-
 ```{r}
@@
 
+\newpage
+
 # Model {#sec-model}
 
-Modeling traffic dynamics poses unique challenges, particularly with count data that often exhibit over-dispersion—where the variance exceeds the mean—making conventional regression models like Poisson unsuitable.
-To address this, this study employs a negative binomial regression model in order to handle over-dispersion and more aligned with the distribution characteristics of traffic counts.
+Modelling traffic dynamics poses unique challenges, particularly with count data that often exhibit over-dispersion—where the variance exceeds the mean—making conventional regression models like Poisson unsuitable.
+To address this, this study employs a negative binomial regression model to handle over-dispersion and is more aligned with the distribution characteristics of traffic counts.
 
@@
 
-This modeling approach not only provides insights into the immediate effects of transit policy changes but also establishes a methodological framework for future studies examining the longitudinal impact of urban transit interventions.
+This modelling approach not only provides insights into the immediate effects of transit policy changes but also establishes a methodological framework for future studies examining the longitudinal impact of urban transit interventions.
 
@@
 # Visualizing residuals to identify patterns
-#par(mfrow = c(1, 2))
-#plot(residuals(model_before), type = "p", main = "Residuals of Model Before Improvement")
-#plot(residuals(model_after), type = "p", main = "Residuals of Model After Improvement")
 
@@
 
+\newpage
+
 ```{r}
+#| label: tbl-modelbeforesummaries
+#| tbl-cap: "Model Summary of Traffic Patterns Before Improvement on the 508 Lakeshore Route"
 #| echo: FALSE
@@
   mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))
-```
-
-```{r}
-#| label: tbl-modelbeforesummaries
-#| tbl-cap: "Model Summary of Traffic Patterns Before Improvement on the 508 Lakeshore Route"
-#| echo: FALSE
-#| warning: FALSE
-#| message: FALSE
 
@@
 
+# Read and prepare data
+analysis_data$count_date <- as.Date(analysis_data$count_date)
+
+# Scaling predictor variables for analysis
+analysis_data <- analysis_data %>%
+  mutate(
+    daily_bus = scale(daily_bus, center = TRUE, scale = TRUE),
+    daily_peds = scale(daily_peds, center = TRUE, scale = TRUE),
+    daily_bike = scale(daily_bike, center = TRUE, scale = TRUE)
+  )
+
+# Splitting the dataset into before and after a specific date for comparison
+start_before <- as.Date("2011-12-31")
+end_before <- as.Date("2014-12-31")
+start_after <- as.Date("2018-11-01")
+end_after <- as.Date("2020-02-29")
+
+before_improvement <- subset(analysis_data, count_date >= start_before & count_date <= end_before)
+after_improvement <- subset(analysis_data, count_date >= start_after & count_date <= end_after)
+
+analysis_data <- analysis_data %>%
+  mutate(time_trend = as.numeric(count_date - min(count_date, na.rm = TRUE)))
+
 # Building negative binomial regression models for before and after improvement
@@
 The increase in the intercept to 12.283 suggests an elevated baseline of car traffic, potentially as a result of external influences not accounted for in the model.
-The coefficient for daily bus traffic experiences a considerable reduction to 0.096, signaling a weakened relationship between bus and car traffic in the wake of the transit improvements.
-This could indicate a disconnect of these modes of transport, possibly due to more passengers opting for the improved public transit system over personal vehicles.
+The coefficient for daily bus traffic experiences a considerable reduction to 0.096, signalling a weakened relationship between bus and car traffic in the wake of the transit improvements.
+This could indicate a disconnect between these modes of transport, possibly due to more passengers opting for the improved public transit system over personal vehicles.
 The time trend coefficient inverses to -0.001033, denoting a statistically significant decline in car traffic over time post-improvement, showcasing the potential efficacy of the public transit enhancements in reducing car traffic.
@@
 
-This study’s analysis, centered on the Route 508 Lake Shore streetcar enhancements in Toronto, demonstrates a statistically significant divergence of bus traffic from car congestion trends, suggesting that improvements to public transit infrastructure can effectively reduce car congestion in urban areas.
-Specifically a decrease in the coefficient for `daily_bus` from 0.261 to 0.096 post-improvement, indicates a notable shift in travel behaviour.
+This study’s analysis, centred on the Route 508 Lake Shore streetcar enhancements in Toronto, demonstrates a statistically significant divergence of bus traffic from car congestion trends, suggesting that improvements to public transit infrastructure can effectively reduce car congestion in urban areas.
+Specifically, a decrease in the coefficient for `daily_bus` from 0.261 to 0.096 post-improvement, indicates a notable shift in travel behaviour.
 
@@
 
-Moreover, the benefits of improved public transit does not stop at decreased car congestion.
-Insights from the @APTA2023, which illustrate that public transit investments yield substantial congestion relief and economic benefits.
-Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested .
-While on the individual level, households in America can save more than \$13,000 per year by using public transit and owning one less car.
+Moreover, the benefits of improved public transit do not stop at decreased car congestion.
+Insights from the @APTA2023, illustrate that public transit investments yield substantial congestion relief and economic benefits.
+Specifically, every dollar spent on public transportation generates 5 dollars in economic return and an associated creation of approximately 50,000 jobs for every billion dollars invested.
+On the individual level, households in America can save more than \$13,000 per year by using public transit and owning one less car.
 
@@
 
-Public transit systems supports not only the movement of people from place to place but also acts a vital channel for economic empowerment and social integration.
+Public transit systems support not only the movement of people from place to place but also act as a vital channel for economic empowerment and social integration.
 Access to jobs, education, healthcare, and cultural institutions are fundamental rights, and equitable public transit policies are crucial to ensuring that these services are available to all individuals, regardless of their socioeconomic status.
@@
 By facilitating or hindering access to critical resources, transportation policies not only impact economic outcomes but also health and environmental conditions.
-This is evident as inadequate or unaffordable public transit disproportionately affects low-income populations, limiting their ability to reach jobs, healthcare, and thereby affecting their overall economic and health status.
+This is evident as inadequate or unaffordable public transit disproportionately affects low-income populations, limiting their ability to reach jobs and healthcare, and thereby affecting their overall economic and health status.
 Additionally, prioritization of car travel over public transit can lead to increased emissions, affecting urban air quality and contributing to environmental degradation.
@@
 These elements have the potential to influence public transit usage patterns and public perception significantly.
-One of the study's primary limitations is its narrow analytical lens, concentrated on traffic volume data, which potentially overlooks the other factors that might influence traffic patterns.
+One of the study's primary limitations is its narrow analytical lens, concentrating on traffic volume data, which potentially overlooks the other factors that might influence traffic patterns.
 For instance, changes in the local economy or shifts in population density.

### 2024-04-19 scripts/02-data_cleaning.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans the raw traffic counts recorded into a more manageable dataset 
 # Author: Maria Mangru
-# Date: March 2024
+# Date: 18th April, 2024
 # Contact:maria.mangru@mail.utoronto.ca
 # License: MIT
 
@@ -103,4 +103,5 @@ daily_traffic <- daily_traffic %>%
 write.csv(daily_traffic, here("data", "analysis_data", "cleaned-data.csv"), row.names = FALSE)
 
 # Saving as Parquet
-write_parquet(daily_traffic, here("data", "analysis_data", "cleaned-data.parquet"))
+write_parquet(x = daily_traffic, sink = here("data", "analysis_data", "daily_traffic.parquet"))
+

### 2024-04-19 scripts/03-exploring_cleaned_data.R

@@ -51,7 +51,7 @@ line_graph_car <- analysis_data %>%
 line_graph_car
 
 # Save car traffic plot
-ggsave("./other/outputs/car_traffic_over_time.png", plot = line_graph_car, width = 12, height = 6, units = "in")
+ggsave("./other/outputs/car_traffic_over_time.png", plot = line_graph_car, width = 12, height = 6, units = "in", create.dir = TRUE)
 
 # Visualizing daily bus traffic over time
 line_graph_buses <- analysis_data %>%
@@ -66,7 +66,7 @@ line_graph_buses <- analysis_data %>%
 line_graph_buses
 
 # Save bus traffic plot
-ggsave("./other/outputs/bus_traffic_over_time.png", plot = line_graph_buses, width = 12, height = 6, units = "in")
+ggsave("./other/outputs/bus_traffic_over_time.png", plot = line_graph_buses, width = 12, height = 6, units = "in", create.dir = TRUE)
 
 # Visualizing daily pedestrian traffic over time
 line_graph_pedestrians <- analysis_data %>%
@@ -81,7 +81,7 @@ line_graph_pedestrians <- analysis_data %>%
 line_graph_pedestrians
 
 # Save pedestrian traffic plot
-ggsave("./other/outputs/pedestrian_traffic_over_time.png", plot = line_graph_pedestrians, width = 12, height = 6, units = "in")
+ggsave("./other/outputs/pedestrian_traffic_over_time.png", plot = line_graph_pedestrians, width = 12, height = 6, units = "in", create.dir = TRUE)
 
 # Visualizing daily bike traffic over time
 line_graph_cyclists <- analysis_data %>%
@@ -96,7 +96,7 @@ line_graph_cyclists <- analysis_data %>%
 line_graph_cyclists
 
 # Save cyclist traffic plot
-ggsave("./other/outputs/cyclist_traffic_over_time.png", plot = line_graph_cyclists, width = 12, height = 6, units = "in")
+ggsave("./other/outputs/cyclist_traffic_over_time.png", plot = line_graph_cyclists, width = 12, height = 6, units = "in", create.dir = TRUE)
 
 
 
@@ -116,7 +116,8 @@ time_trend_graph_car <- analysis_data %>%
 time_trend_graph_car
 
 # Save the car traffic plot with time trend
[…]

### 2024-04-19 scripts/04-test_data.R

@@ -84,39 +84,33 @@ test_that("Dates are within the filtered range of 2014 to 2019", {
 })
 
 
-
 #### Tests for exploring cleaned data ####
 source(here("scripts", "03-exploring_cleaned_data.R"))
 
 # Test summary statistics generation
 test_that("Summary statistics are generated correctly", {
   expect_false(is.null(summary_stats))
-  expect_true("mean_daily_cars" %in% names(summary_stats))
-  expect_true("median_daily_cars" %in% names(summary_stats))
-  expect_true("mean_daily_bus" %in% names(summary_stats))
-  expect_true("median_daily_bus" %in% names(summary_stats))
-  expect_true("mean_daily_peds" %in% names(summary_stats))
-  expect_true("median_daily_peds" %in% names(summary_stats))
-  expect_true("mean_daily_bike" %in% names(summary_stats))
-  expect_true("median_daily_bike" %in% names(summary_stats))
+  expect_true("Mean_Cars" %in% names(summary_stats))
+  expect_true("Median_Cars" %in% names(summary_stats))
+  expect_true("Mean_Buses" %in% names(summary_stats))
+  expect_true("Median_Buses" %in% names(summary_stats))
+  expect_true("Mean_Pedestrians" %in% names(summary_stats))
+  expect_true("Median_Pedestrians" %in% names(summary_stats))
+  expect_true("Mean_Bicycles" %in% names(summary_stats))
+  expect_true("Median_Bicycles" %in% names(summary_stats))
 })
+
 
 # Test that plot data is structured correctly
 test_that("Data for visualizations is prepared correctly", {
-  expect_true("count_date" %in% names(cleaned_data))
-  expect_true("daily_cars" %in% names(cleaned_data))
-  expect_true("daily_bus" %in% names(cleaned_data))
-  expect_true("daily_peds" %in% names(cleaned_data))
-  expect_true("daily_bike" %in% names(cleaned_data))
+  expect_true("count_date" %in% names(analysis_data))
+  expect_true("daily_cars" %in% names(analysis_data))
+  expect_true("daily_bus" %in% names(analysis_data))
+  expect_true("daily_peds" %in% names(analysis_data))
+  expect_true("daily_bike" %in% names(analysis_data))
 })
 
-# Test that plot objects are created
-test_that("Plot objects are created", {
-  expect_false(is.null(line_graph_car))
-  expect_false(is.null(line_graph_buses))
[…]

### 2024-04-19 scripts/05-model.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Creates negative binomial model 
 # Author: Maria Mangru
-# Date: March 2024
+# Date: 18th April, 2024
 # Contact:maria.mangru@mail.utoronto.ca
 # License: MIT
 
@@ -53,7 +53,7 @@ summary(model_after)
 
 
 
-# Graphical representation of the model results for the 'before' improvement model
+# Graphical representation of the model results for the 'before' and `after` improvement model
 
 # For model_before
 before_fitted <- data.frame(observed = before_improvement$daily_cars, fitted = fitted(model_before))
@@ -76,10 +76,6 @@ ggplot(after_fitted, aes(x = fitted, y = observed)) +
        y = "Observed Values") +
   theme_minimal()
 after_fitted
-
-# Save the plot
-ggsave(here("other", "outputs", "before_plot.png"), plot = time_trend_graph_car, width = 12, height = 6, units = "in")
-
 
 
 #### Save model ####
